Electroconvulsive stimulation drives cortical spreading depression dependent immediate early gene expression in mice

Version 1

Abstract

Electroconvulsive therapy (ECT) is a highly effective treatment for several psychiatric disorders, though its biological mechanisms remain unclear. Its therapeutic action has traditionally been attributed to the generalized seizure ECT induces. However, this view is challenged by the recent finding that electroconvulsive stimulation (ECS) can trigger a cortical spreading depression (CSD). Because CSD triggers massive intracellular molecular changes, we hypothesized that it could be a key mediator of ECT’s therapeutic, plasticity-inducing effects. We observed similar neuronal oscillations following ECS in mice and patients undergoing ECT. We show that CSD drives increased expression of the immediate early gene Fos, a key marker of neuronal plasticity, and is associated with factors that predict positive ECT therapeutic outcome. Our results suggest that the therapeutic efficacy of ECT may be mediated by CSD. This challenges the seizure-centric model and implies that CSD, a currently unmonitored neurophysiological event, may serve as a more relevant biomarker for predicting and optimizing therapeutic outcomes of ECT.

Introduction

Electroconvulsive therapy (ECT) is one of the most potent and rapidly acting interventions in clinical psychiatry. It demonstrates substantial efficacy in treating severe and often treatment-resistant conditions, including major depressive disorder (Trifu et al., 2021), bipolar disorders (Tor et al., 2021), catatonia (Perugi et al., 2017), mania (Martin et al., 2025) and schizophrenia (Grover et al., 2019). ECT frequently achieves positive therapeutic outcomes in terms of remission rates and speed of response that surpass standard pharmacotherapies, making it indispensable in urgent clinical situations such as high suicide risk or severe catatonia (Baghai and Möller, 2008). ECT is also one of the few treatment options available to drug-resistant patients (Haddad and Correll, 2018). Despite its established clinical utility and decades of refinement aimed at enhancing safety and tolerability, the fundamental neurobiological mechanisms driving ECT’s therapeutic effects remain poorly understood.

There is a multitude of biomarkers affected by ECT that could underlie its mode of action as a psychiatric treatment (Singh and Kar, 2017). These include, non-exhaustively, the modulation of various neurotransmitter systems (Baldinger et al., 2014), alterations in regional cerebral blood flow (Takano et al., 2011) and metabolism (McCormick et al., 2007), changes in functional brain network connectivity (Ulrich et al., 2025), and the promotion of neuroplasticity (Bouckaert et al., 2014; Chen et al., 2018; Joshi et al., 2016). Understanding how these effects contribute to therapeutic outcomes is a significant challenge. We suspect that a critical gap in our knowledge lies in the early cascade of events triggered by an ECT treatment.

ECT, as it is used today, is targeted at the induction of a controlled seizure under general anesthesia and muscle relaxation through an electroconvulsive stimulation (ECS). Import to note here is that the term seizure likely refers to a different neuronal event in the context of ECT and epilepsy. In modern practice, ECS triggered seizures are highly synchronized brain-wide oscillations, which typically follow a three-phase progression (Brumback and Staton, 1982) consisting of an immediate 14-22 Hz activity (phase I), which evolves into arhythmic poly-spike patterns (phase II) and is then rapidly, within seconds, followed by a sustained low-frequency oscillation at 2-4 Hz (phase III). Phase III oscillations often occur without any ictal spikes that are characteristic of epileptic seizures (Brumback and Staton, 1982) and share limited similarities only with the late phase of seizures observed in epilepsy (Pottkämper et al., 2021). The phase III oscillation is referred to as seizure in the ECT literature. Illustrating the difference between what is referred to as a seizure in the ECT literature and what is referred to as a seizure in the epilepsy literature is also the fact that the post-seizure neuronal silencing of activity is a marker of treatment success in ECS, while it is associated with sudden unexpected death in epilepsy (Krystal et al., 1995; Sowers et al., 2013), and the finding that antiepileptics do not block ECS-triggered seizures (Cinderella et al., 2022). To try to disambiguate this, we will refer to the ECS induced seizure as phase III oscillation in the description of our results, but consistent with previous usage as seizure when talking about prior ECT work.

Historically, the generalized ECS induced seizure was considered to be the essential therapeutic component of the treatment (Accornero, 1988; Ulett et al., 1956). However, the occurrence and properties of seizures are only partially predictive of therapeutic outcomes (Perera et al., 2004; Sackeim et al., 1991). Similarly, the clinical efficacy of alternative stimulation methods, such as low amplitude ECT or magnetic seizure therapy, demonstrates that robust outcomes can be achieved with lower field strengths (Peterchev et al., 2015). This discrepancy suggests that electric field strength is unlikely to be the primary determinant of therapeutic success (Deng et al., 2024), indicating that ECT’s efficacy hinges on biological mechanisms beyond the mere presence of a seizure or the magnitude of the electrical stimulus.

A secondary neurophysiological event frequently triggered by seizures is cortical spreading depression (CSD). It has been speculated that CSD is an endogenous mechanism of the brain used for seizure termination (Dreier et al., 2012; Kelly et al., 1999). A CSD can also be triggered by direct electrical stimulation of the cortex (Fregni et al., 2007; Leao, 1944), and, consistent with this, recent evidence revealed the existence of CSD-like calcium events following ECS in mice and likely following ECT in patients (Rosenthal et al., 2025). This discovery necessitates a critical re-evaluation of the traditional seizure-centric view of the effects of ECT. The profound neurophysiological disruption inherent to CSD may potentially act as a trigger for several of the neurobiological alterations driven by ECT.

CSD is characterized as a slowly propagating wave (typically 2–5 mm/min) of near-complete, and sustained depolarization affecting both neurons and glial cells (Seidel et al., 2016). This wave of intense depolarization is invariably followed by a period of marked suppression of spontaneous and evoked electrophysiological activity that lasts several minutes and can persist for up to an hour (Sawant-Pokam et al., 2017). The wavefront of the CSD is a massive transmembrane ionic flux (Somjen, 2001) that is accompanied by cellular swelling due to water movement, blood flow changes resulting in local hypoxia, and structural changes to dendritic processes (Takano et al., 2007). A CSD can be triggered by a large efflux of potassium ions into the extracellular space (Brinley et al., 1960) that causes a positive feedback loop via glutamate release, which propagates the wavefront through the activation of NMDA receptors (Zhou et al., 2013). Concurrently, there is an influx of sodium, chloride and calcium ions into cells (Pietrobon and Moskowitz, 2014). Disturbed calcium signaling is associated with a host of neuropsychiatric disorders treated by ECT (Nanou and Catterall, 2018), which makes CSD driven calcium influxes a likely key contributor to neuronal plasticity effects driven by ECT (Dehbandi et al., 2008; Wernsmann et al., 2006).

Here, we sought to characterize the specific calcium dynamics during ECS-induced CSD, hypothesizing they might underlie the plasticity-relevant effect of ECT. Consistent with previous work (Rosenthal et al., 2025), we observed a slow travelling calcium event that appeared to be a CSD. We found that the characteristics of this calcium event were consistent with being a CSD, and that it accounted for the ECS driven Fos expression, a marker of neuronal plasticity in the cortex (Mahringer et al., 2022, 2019). Furthermore, we show that EEG metrics known to predict positive therapeutic outcomes of ECT are correlated with the occurrence of CSD in mice. Our findings support the hypothesis that CSD, rather than the seizure itself, is the primary driver of ECS driven plasticity and, by extension, the therapeutic benefits of ECT.

Results

ECS triggers similar neuronal oscillations in both mice and humans

We first asked how well ECS in mice recapitulates EEG signatures of ECS in humans. To do this, we measured neuronal responses following ECS in mouse dorsal cortex, using concurrent widefield calcium imaging and EEG recordings (Methods). We expressed a genetically encoded calcium indicator brain-wide by retroorbital deposit of an AAV-PHP.eB-Ef1α-GCaMP6s-WPRE and performed concurrent EEG recordings using a semitransparent 30-channel EEG electrode (Figure 1A). During ECS, mice were anesthetized with isoflurane (2%) and myorelaxed with succinylcholine chloride (1mg/kg). ECS was delivered through biauricular electrodes using a commercial ECS unit. Stimulation parameters were set within a range that reliably evoked tonic-clonic seizures in mice (Methods).

Electroconvulsive stimulation (ECS) triggers similar slow neuronal oscillations in mice and humans.

(A)Top: Schematic of mouse ECS setup. We performed concurrent widefield calcium imaging and EEG in the mouse dorsal cortex during biauricular ECS. Bottom: Layout of the EEG probe. (B)Spectrograms of two widefield calcium imaging ROIs surrounding EEG electrodes L12 and R12 (as highlighted in A). Duration of the ECS is marked by a black bar. (C)Spectrograms of the EEG recorded concurrently with the widefield calcium data in C. Time during which EEG is not recorded due to a stimulation artifact is marked by gray shading. (D)Schematic of patient ECS setup. EEG recordings were performed using two frontal electrodes after bifrontal ECS in patients. (E)Spectrograms of EEG recorded following ECT in a patient. Patient EEG recordings are started following ECS and terminated when the oscillation stops. Gray shading marks times without EEG recording. (F)Top: Oscillations observed in early, mid, and late phases after ECS in an example mouse widefield calcium ROI (blue) and in the concurrently recorded EEG (green). Bottom: Oscillations observed in an example patient EEG recording. (G)Oscillation duration for mouse and patient EEG recordings. The central box represents the interquartile range (IQR), spanning from the first (Q1) to the third (Q3) quartiles. Whiskers extend to data points not considered outliers (Methods). A central horizontal line indicates the median and an open circle the mean of the data distribution. (H)Power-weighted mean frequency at peak and offset of oscillations for mouse widefield, mouse EEG and patient EEG recordings. Here and elsewhere: n.s.: not significant, *: p < 0.05, **: p < 0.01, ***: p < 0.001. See Table S1 for all information on statistical testing.

Across the parameter range used, ECS triggered an immediate and brief calcium response across the entire dorsal cortex (Figure S1A). The response onset occurred with a latency of less than 10 ms (within the first frame after stimulation at 100 Hz acquisition), and calcium activity continued to increase until the end of the ECS (Figure S1B). In the EEG data (Figure S1C) we were blind to this direct response, as we disconnected the EEG from the data acquisition system during ECS to prevent electrical arcing between electrodes (Methods). We speculate that the direct ECS response observed in calcium imaging is driven by spiking activity in the cortical neurons, given that the electrical field generated by the ECS is several times higher than the neuronal activation threshold (Deng et al., 2011). Following the direct calcium response to ECS, mean calcium activity decreased and returned to baseline only slowly (Figure S1B).

The combined median duration of phases I and II is 9 s (Hogan et al., 2019). Consistent with this, we observed a phase III slow oscillation in the delta to low theta range (1 to 5 Hz) in mice that emerged about 10 s after the ECS, both in calcium activity and EEG recordings (Figure 1B-C, S2). This oscillation was strikingly similar to that observed in patients undergoing ECT (Figure 1D-E) (Stuiver et al., 2026), and was present in the EEG recording of all patients analyzed here. Given that ECT stimulation parameters are chosen to increase the likelihood of phase III oscillations, which in turn are used as a biomarker for treatment outcome (Sackeim et al., 1991), this is perhaps not surprising.

The oscillation followed a similar pattern in both mice and patients and typically exhibited a strong sinusoidal shape in the early stage that desynchronized over time (Figure 1F). In mice, the oscillations lasted 42.3 ± 11.9 s (mean ± std), and we found no evidence of a difference in oscillation duration between mice and patients (Figure 1G). These values are consistent with previous estimates of the duration of phase III oscillations in patients (Hogan et al., 2019). Interestingly, the frequencies of the oscillations were also very similar in mice and patient EEG, and both tended to decrease with time (Figure 1H). The estimate of the oscillation frequency in the mouse calcium imaging data was systematically higher relative to the concurrent EEG data (Figure 1H) but exhibited the same decreasing trend. We suspect this difference is driven by differential contributions to the two signals from non-oscillation related neuronal activity and noise sources.

ECS can trigger a travelling calcium event

We found that ECS could also trigger a travelling calcium event, as previously described (Rosenthal et al., 2025). To monitor the propagation of these events, we recorded the calcium activity across the entire dorsal cortex for 3 to 5 minutes following ECS (Figure 2A). They manifested as a significant increase in calcium fluorescence that propagated across the dorsal cortex (Figure 2B) at a speed of 0.083 ± 0.009 mm/s (mean ± std) (Figure 2C), consistent with previously reported CSD propagation speeds in mice (Enger et al., 2015). Calcium events exhibited maximum fluorescence intensities that exceeded even those observed during the direct ECS response (Figure 2D). Once initiated, these events spread across an entire dorsal cortex hemisphere over the course of several minutes (Figure 2E). In line with the refractory dynamics of CSD, the event exhibited a non-repeating trajectory, traversing each part of dorsal cortex only once. The lateralization of the events exhibited a certain stochasticity and calcium events were not always symmetrically present in both hemispheres. We frequently observed cases in which the events were unilateral and constrained to only left or right cortical hemispheres (Figure 2F-G).

ECS can trigger a travelling calcium event.

(A)Left: Schematic of mouse ECS setup without concurrent EEG recordings. Right: Widefield recordings of mouse dorsal cortex were parcellated into 818 ROIs, each of which was assigned to a cortical region according to the Allen Brain Atlas taxonomy. The ROIs highlighted correspond to the calcium activity shown in B and E. (B)Example of calcium activity in ROIs highlighted in A during the traversal of a calcium event. (C)Propagation speed of calcium events (Methods). Each data point is the median speed across one calcium event. (D)Maximum calcium activity throughout the dorsal cortex during pre-ECS, direct ECS response and calcium event periods. Two direct ECS responses are outside the plot range and not shown (1707 % and 2374 % ΔF/F0). (E)Left: Low-passed (< 0.5 Hz) calcium activity during an example ECS session triggering calcium event over both hemispheres. Each panel shows the average calcium activity over a 10 ms window starting at the time indicated above the panel. Right: Isochrones (spaced at two seconds increments) indicating the time at which each region reaches 200% of its baseline fluorescence. (F)As in E, but for a unilateral calcium event over the right hemisphere following a 150 s delay post-ECS. (G)As in E, but for a unilateral calcium event over the left hemisphere.

In about one third of cases, calcium events appeared bilaterally, while in the remaining two thirds the events were confined to one hemisphere only (Figure 3A, Video S1). The majority of unilateral events occurred in the right hemisphere, and across all events, most first appeared on the right edge of the cranial window (Figure 3B). This right hemisphere bias was likely driven by the ECS polarity, as we always placed the positive pole of stimulation on the right ear (Figure 1A). Calcium events were the only type of response to ECS that occurred unilaterally (Figure 3C). While all calcium events co-occurred with oscillations, just over half of the oscillations were accompanied by a calcium event (Figure 3D). The likelihood of triggering a calcium event has been shown to be a function of the stimulation current and frequency used of the ECS (Rosenthal et al., 2025). To investigate how the stimulation parameters control the appearance of oscillations and calcium events, we tested stimulation frequencies between 25 and 100 Hz and stimulation currents between 15 and 75 mA for a total charge delivery of between 0.187 and 3.750 mC. With most combinations of parameters, we observed a non-zero likelihood of triggering a phase III oscillation (Figure 3E). The threshold for triggering calcium events was systematically higher and appeared to be a function of both stimulation frequency and current (Figure 3E). Based on this, we hypothesized that total charge delivery might be a good predictor of the occurrence of calcium events. A linear regression of total stimulation charge against the likelihood of triggering a calcium event showed a significant correlation (Figure 3F). Stimulation frequency and stimulation current also exhibited a positive but nonsignificant correlation with the likelihood of triggering a calcium event (Figure S3A). A multivariate linear regression confirmed that the interaction term of current and frequency was significant and non-zero (0.015 % / (mA Hz)). We found a similar relationship between the total stimulation charge and the probability of triggering a phase III oscillation (Figure 3G, S3B). Consistent with the hypothesis that the calcium event exhibits all-or-none dynamics, we found no evidence of changes in propagation speed, total duration, or delay to onset of calcium events as a function of total stimulation charge (Figure S3C-E). We also observed that the total ECS charge correlated with the intensity of the direct ECS response (Figure S4A). This direct response, in turn, predicted the hemisphere of the calcium event in 75% of the unilateral cases (Figure S4B), supporting our hypothesis that the ECS stimulation parameters can control the occurrence of calcium events.

Calcium events are triggered preferentially by higher total ECS charge.

(A)Probability distribution of the laterality of calcium events. (B)Spatial distribution of calcium events detection sites. For bilateral calcium events, only the hemisphere where the calcium event was detected earliest is considered as the detection site. (C)Fraction of stimulation responses, oscillations and calcium events that were bilateral. (D)Probability of oscillations being observed if a calcium event is triggered (P(osc. | calcium event)), of a calcium event being observed if an oscillation is triggered (P(calcium event | osc.)), and of neither being observed (P(neither)). (E)ECS parameter map of the likelihood of triggering a calcium event (top) or an oscillation (bottom). (F)Probability of triggering a calcium event as a function of total ECS stimulation charge. The solid line represents the linear fit of the data; the shaded region represents 95% confidence interval on the fit. (G)As in E, but for the probability of triggering an oscillation.

The calcium event exhibits hallmark features of CSD

The mode of propagation, speed and fluorescence intensity support the hypothesis that the travelling calcium event is a CSD. This has already been speculated previously (Rosenthal et al., 2025), and makes a number of directly testable predictions. First, a CSD changes the local ionic homeostasis and is accompanied by hemodynamic constriction (Takano et al., 2011). To test for hemodynamic effects triggered by ECS, we performed widefield imaging experiments in a cohort of mice in which we expressed enhanced green fluorescent protein using an AAV (EGFP,AAV-PHP.eB-Ef1α-EGFP-WPRE). In widefield imaging, the EGFP fluorescence signal depends on pH and overlying blood volume. The latter is a phenomenon called hemodynamic occlusion, whereby decreased blood vessel diameter increases the EGFP signal (Ma et al., 2016; Yogesh et al., 2025). During the ECS, we observed an immediate increase in the EGFP signal that lasted for a couple of seconds and is consistent with acute vasoconstriction (Figure S5A). We also observed travelling signals in the EGFP recordings with spatial and temporal characteristics that resembled the calcium event (Figure S5B). The amplitude of these events was smaller than that observed in calcium imaging (Figure S5C), but they propagated at a speed that did not differ significantly from that observed in calcium events (0.096 ± 0.02 mm/s, Figure S5D). This would be consistent with either vasoconstriction or an increase in intracellular pH. CSD is associated with vasoconstriction (Takano et al., 2007), but typically only a brief alkalization over the course of seconds followed by a more prolonged acidification of the intracellular domain over the course of minutes (Sun et al., 2011). However, given that an estimated alkalization of less than 0.5 in pH, would translate to less than a 10% ΔF/F0 change in EGFP fluorescence (Kneen et al., 1998), we suspect most of the change we observe is driven by hemodynamics.

A second feature of CSD is the propagation through extracellular diffusion (Enger et al., 2015), rather than through electrical activity. Increases in intracellular calcium levels in neurons are typically associated with increases in spiking activity. A CSD, however, is associated only with brief increases in spiking activity at the onset (Herreras et al., 1994; Nasretdinov et al., 2023). To visualize the propagation of the calcium event in the tissue, we used two-photon microscopy to observe the wavefront at cellular resolution. We expressed a genetically encoded calcium indicator (AAV2/9-Ef1α-GCaMP6f-WPRE) in visual cortex and measured responses to ECS in layer 2/3 neurons (Figure 4A, Video S2). Consistent with a diffusion-limited process, we found that the calcium event was a sharp wavefront of significant increase in fluorescence, propagating across the entire field of view (Figure 4B) at a speed of 0.079 ± 0.0174 mm/s (mean ± std). This aligns with our speed estimates based on widefield calcium imaging. Combined with the lack of propagation along inter-hemispheric or other known anatomical projections (Figure 2E-G), this supports the idea that the calcium events are propagating through a diffusion-limited process, rather than synaptic activity, consistent with a CSD.

Calcium events first spread in supragranular layers and elicit long-lasting calcium increase in cells.

(A)Schematic of the experimental setup. We performed two-photon (2p) calcium imaging in the right primary visual cortex during ECS. (B)2p calcium activity during an example ECS session triggering a calcium event. Time of maximum (TOM) is shown on the right. Each panel shows the average calcium fluorescence over a 500 ms window starting at the time indicated above the panel. Gray inset highlights two example somata whose calcium signals are shown in G. (C)Top: Schematic of the micro-prism implanted below the imaging window to allow for simultaneous recording in all cortical layers (see Methods). Bottom: Widefield imaging of the implant 18 seconds after ECS, confirming the prism does not obstruct the calcium event propagation. A gray arrow indicates the orientation of the prism. (D)As in B, but imaged through a prism implant. Cortical layers 3, 4, and 5a (L3, L4, L5a) are defined based on soma densities (Methods). (E)Calcium fluorescence of superficial (blue) and deep (green) layers somata during calcium events, aligned to the time of maximum of all neuropil (Methods). Lines represent the hierarchical bootstrap estimates of the mean value for each time bin. Here and elsewhere, shading around the mean is one standard deviation of the bootstrap distribution at each time bin. Response curves are compared for each time bin: in the horizontal bars above the plot, black marks time bins where p<0.05 and gray mark time bins where p>0.05. (F)Time of 50% of maximum activity for superficial and deep layers somata during calcium events. (G)Top: Calcium fluorescence from the two example somata in B. Bottom: Calcium fluorescence from the neuropil surrounding these somata. Fitted slopes during the 60 seconds following calcium event onset are shown as dashed lines. (H)Calcium fluorescence of somata (dark gray) and neuropil (purple) in all layers during calcium events. Each soma is aligned to the time of maximum of the neuropil. (I)Distribution of the fitted slopes for cells and neuropil during the 60 seconds following calcium events. (J)Scatter plot of fitted slopes and correlation of cells with their neuropil. Top and right histograms are marginal distributions over slopes and correlations, respectively.

A third hallmark feature of CSD is its preferential propagation through superficial cortical layers (Stafstrom, 2020; Zakharov et al., 2019). To investigate this, we implanted a micro-prism in the cortex, enabling simultaneous imaging across all cortical layers (Figure 4C, Video S3). We found that the calcium event first spread in superficial layers 2/3 and 4, and only later reached deeper layers 5 and 6 (Figure 4D). Calcium event response in superficial layers preceded those in deep layers by 6.9 ± 1.3 s (mean ± std) (Figure 4E-F). While both somata and neuropil showed a sharp calcium fluorescence increase at the wave front, calcium decay kinetics were more heterogeneous. Calcium fluorescence in the neuropil gradually returned to baseline over the course of about 100 s. Most somata followed this time course, but a subset of them exhibited a progressive increase in fluorescence (Figure 4G), likely reflecting a slow accumulation of somatic calcium. This resulted in a population average fluorescence decrease that was significantly faster in the neuropil than in the somata (Figure 4H-I). The somata with increasing post-calcium event fluorescence made up about 35% of the total somatic population (Figure 4J). This heterogeneous response pattern is a characteristic of CSD, and is hypothesized to be linked to the variable capacity of individual neurons to recover from CSD-induced oxidative stress (Enger et al., 2015). Overall, these results strongly support the hypothesis that the ECS triggered calcium events are indeed CSD.

Calcium events, not ECS, drives Fos expression

If those calcium events are CSD, then they should cause neuronal plasticity (Faraguna et al., 2010; Li et al., 2012). To begin testing this hypothesis, we quantified the ECS driven expression of the Fos protein, the product of an immediate early gene that is a marker of functional plasticity in cortex and hippocampus (Mahringer et al., 2022, 2019). We again performed widefield calcium imaging of ECS responses and allowed one hour of recovery post-ECS for Fos expression (Methods). Brain tissue was then processed for Fos immunohistochemistry. We found that ECS drove strong Fos expression in a variety of brain areas, including the cortex (Figure 5A). To determine whether it was the direct electrical response to the ECS or the calcium event that drove Fos expression, we again took advantage of the occurrence of unilateral calcium events (Figure 5B). In these cases, we could directly compare Fos expression in the hemisphere that exhibited a calcium event to that in the hemisphere that did not. Remarkably, hemispheres without a calcium event showed Fos levels not statistically different from sham-treated mice, and the ECS driven increase in Fos expression could be fully explained by the calcium event (Figure 5C). This supports the hypothesis that calcium events, which are likely CSD, are the primary driver of neuronal plasticity following ECS.

The ECS triggered increase in Fos expression in cortex is driven by the calcium event.

(A)Example of coronal brain sections stained for Fos from three sham ECS (top) and three ECS-treated (bottom) mice. (B)Maximum calcium fluorescence during direct ECS response (first row), the amplitude of the ECS induced oscillation (second row), and the peak fluorescence to illustrate the presence or absence of a calcium event (third row). From left to right, data from three example mice that had unilateral left, unilateral right, and bilateral calcium events, respectively. The bottom row shows corresponding coronal sections from the same brains, stained for Fos expression. (C)Quantification of the Fos throughout the entire cortex, expressed as a fold change for an entire hemisphere normalized to sham mice (see Methods). Paired hemispheres (experiments in which ECS triggered a bilateral direct ECS response and oscillation followed by a unilateral calcium event) are linked by a line. Thus, the calcium event, not the direct ECS response or the phase III oscillation, is the correlate of Fos expression.

The occurrence of calcium events correlates with EEG metrics predicting positive ECT therapeutic outcomes

Given that calcium events accounted entirely for the expression of Fos, we speculated that they may also be the driver of the positive therapeutic outcomes of ECT. A first testable prediction of this hypothesis is that EEG responses to ECS that correlate with positive therapeutic outcomes should also correlate with the occurrence of calcium events. While EEG responses are only partially predictive of clinical improvement following ECT (Krystal and Weiner, 1994), there are several EEG metrics that correlate with positive therapeutic outcomes. For the phase III oscillation, an increased amplitude (Perera et al., 2004), a longer duration (Gillving et al., 2024), a higher post oscillation suppression index (Kimball et al., 2009; Nobler et al., 1993), and higher interhemispheric coherence (Grözinger et al., 2013), for example, are indicators of a favorable patient response to ECT (Francis-Taylor et al., 2020).

To address this, we investigated whether these EEG metrics correlated with the occurrence of a calcium event, by comparing EEG data from recordings without and with calcium events (Figure 6A-B). We found that the mid-phase III oscillation amplitude was significantly higher in hemispheres where a calcium event occurred compared to the unaffected contralateral hemisphere or to ECS where neither hemisphere showed a calcium event (Figure 6C). Both the oscillation duration and postictal suppression index (Figure 6D-E) were increased in ECS triggering a calcium event, but we found no evidence that the laterality of the calcium event (ipsi-vs. contralateral) influenced these two metrics. This absence of hemisphere specificity would be consistent with the observation of a tight coupling of onsets and offsets of phase III oscillations across the hemispheres (Brumback and Staton, 1982). In the case of inter-hemispheric coherence of the EEG signals, we found no evidence of an influence of the calcium event (Figure 6F). These results led us to speculate that calcium events may serve as a central mechanistic driver of the therapeutic efficacy of ECT.

Calcium events are associated with ictal EEG metrics that predict clinical ECT efficacy.

(A)Example of an ECS session triggering a direct ECS response followed by a slow oscillation. Data is shown for two example electrodes. A gray transparent square indicates the duration of the oscillation. First row: Calcium imaging spectrogram of the ROI surrounding the electrode. Second row: EEG signal spectrogram of the electrode. Third row: z-scored EEG signal in the 1-5 Hz band to visualize the oscillation amplitude. Fourth row: EEG Hilbert amplitude in the 1-30Hz band, with dashed lines indicating the mean amplitudes during and after the oscillations used to compute the postictal suppression index (PSI). Fifth row: Calcium imaging fluorescence, low-passed below 0.5 Hz. (B)As in A, with identical ECS stimulation parameters, but for an ECS session triggering a direct ECS response followed by a bilateral calcium event. (C)Mid-oscillation amplitude for each electrode during ECS sessions triggering only direct ECS response (gray), ECS sessions triggering a unilateral calcium event over the contralateral hemisphere (blue) and ECS sessions triggering a calcium event on the ipsilateral hemisphere (red, either unilateral or bilateral calcium event). See Methods for the details on the ictal EEG metrics definitions. (D)As in C, but for the oscillation duration. (E)As in C, but for the postictal suppression index. (F)As in C, but for the inter-hemispheric coherence of the EEG signal.

Discussion

Here, we sought to characterize the neuronal response to ECS, focusing on a calcium event that shows hallmark features of a CSD. We demonstrate that these events correlate with prognostic EEG markers of ECT treatment outcomes and induce the expression of Fos, a key indicator of neuronal plasticity. Based on our results, we propose that CSD may be a more relevant biomarker than the phase III oscillation to explain the therapeutic effects of ECT.

Limitations

There are two key limitations that constrain the interpretation of our results, and that will need to be addressed in future research:

We have no direct evidence that patients also have ECT-induced CSD. Patients are typically monitored with two-channel EEG to determine whether ECT has induced a phase III oscillation (Francis-Taylor et al., 2020), but these measurements are currently not sufficient to detect CSD. One simple way to modify the current clinical routine measurements would be to perform direct current (DC) EEG recordings with a higher channel count. CSD results in characteristic DC shifts in the EEG response (Hartings, 2016). While detecting stochastically-occurring CSD signatures in EEG recordings of awake subjects is difficult (Lauritzen et al., 2011), this would likely be facilitated in the context of ECT by the known timing of the triggering stimulus and the elimination of movement artifacts via anesthesia and myorelaxation (Hofmeijer et al., 2018). Alternatively, given that the CSD triggers a strong hemodynamic change (Figure S5) (Hadjikhani et al., 2001), blood-oxygenation level dependent imaging techniques might be able to detect a CSD. There is already evidence of hemodynamic changes induced by ECT in patients (Rosenthal et al., 2025).

Our conclusion that the CSD is the primary driver of plasticity is based on its role in driving Fos expression, an early marker of cortical plasticity (Mahringer et al., 2022, 2019). We cannot rule out the possibility that the seizure induced by ECS drives Fos independent plasticity pathways that also contribute to therapeutic outcomes. Like CSD, seizures are neuronal events characterized by supra-physiological excitability that trigger neuronal plasticity (Scharfman, 2002). While knocking out Fos reduces seizure-induced plasticity (Watanabe et al., 1996), other signaling cascades such as mTOR (Kim et al., 2022), BDNF (Ramnauth et al., 2022) and the immediate early gene Arc (Larsen et al., 2005) can all be triggered by ECS-induced seizures. One way to explore the contribution of these plasticity pathways to treatment effects will be to map the gene expression changes following ECT more broadly.

Does ECT drive CSD in humans?

In addition to the results presented in this paper and that of (Rosenthal et al., 2025), there are three additional lines of evidence that support the hypothesis that ECT triggers CSD in patients:

A significant proportion of ECT patients suffer from a migraine-like headache immediately after the treatment (Mulder and Grootens, 2020). Migraines are thought to be caused by a CSD driven calcium influx that activates Panx1 channels (Karatas et al., 2013).

Confusion and cognitive side effects, with a particular focus on memory and hippocampus-dependent processes, are frequently reported following ECT (Impastato and Almansi, 1942; Porter et al., 2020). In mouse models of epilepsy, postictal confusion has been shown to be driven by hippocampal CSD, rather than by the seizure itself (Mitlasóczki et al., 2025).

CSD causes osmotic changes (Mazel et al., 2002) that could explain the structural imaging changes observed after a single ECT session (Takamiya et al., 2025; Verdijk et al., 2026), which are particularly pronounced in the hippocampus (Bracht et al., 2026; Wilkinson et al., 2017).

Our data show that metrics which predict a positive ECT treatment outcome also correlate with the occurrence of CSD (Figure 6). Given that standard ECT protocols optimize for these metrics as a proxy for seizure quality (Krystal et al., 1995; Sackeim et al., 1991), it is possible that clinical practice has inadvertently optimized ECT to trigger CSD. If CSD accounts for plasticity effects, triggering a CSD in a non-seizure context may be sufficient to elicit therapeutic effects. This is supported by the clinical success of ultra-brief stimulations treatments that do not cause seizures, such as rTMS with accelerated protocols, which achieves treatment efficacy on par with ECT for major depressive disorder (Cole et al., 2022). Similarly, ultrasound stimulation can trigger CSD in mice (Estrada et al., 2021) and functions as a treatment for major depressive disorder (Oh et al., 2024). Much like using a landslide to create ripples in a pond, a generalized seizure may be an unnecessarily burdensome way to trigger the CSD required for treatment-relevant plasticity effects.

CSD as a driver of ECT related plasticity

The massive intracellular calcium influx during CSD (Figure 2, 4) is a prime candidate for driving the molecular changes underlying ECT-induced neuronal plasticity. Calcium is the entry point for a variety of plasticity-related signaling pathways that are disrupted in psychiatric conditions (Nanou and Catterall, 2018). While both physiological activity and seizures can trigger these pathways, CSD stands out due to the sheer magnitude of its calcium surge, driven in part by a near-total release of endoplasmic reticulum stores (Stern et al., 2024). Elevated calcium levels drive the expression of key plasticity markers. For instance, CSD-induced calcium triggers the Fos signaling cascade (Iqbal Chowdhury et al., 2003) and promotes the expression of BDNF (Faraguna et al., 2010). As the primary ligand for the TRKB receptor, BDNF activates a pathway that is a well-established target for both antidepressant (Casarotto et al., 2021) and antipsychotic drugs (Pandya et al., 2013).

While we cannot fully exclude the possibility of a secondary event, such as a unilateral seizure restricted to the CSD hemisphere, we monitored calcium activity in a subset of mice (11/17) until shortly before euthanasia and did not see any seizure-like calcium activity. We also did not observe behavioral signatures of a seizure in any of the mice. But even if secondary events did occur, the first unilateral event we observed was the CSD. This, combined with the fact that the CSD provides a potent biological mechanism to explain the robust neuroplasticity driven by ECT, leads us to conclude that the CSD is the primary treatment relevant event triggered by ECT, and not the phase III oscillation.

Origin of phase III oscillation

In the ECT literature, the phase III oscillation is often referred to as a seizure. We suspect that the key difference to an epileptic seizure is that an initial strong input (either directly from the ECS or indirectly via ECS driven excitation of cortex) to thalamus triggers a 2-4 Hz oscillation (Bal et al., 2000). At the same time, ECS stochastically triggers CSDs. As these move across cortex they silence cortical activity, and what remains in their wake is the oscillatory drive from thalamus. A central thalamic is consistent with the observation that phase III oscillations have zero phase lag across the human cortex (Brumback and Staton, 1982; Huels et al., 2023). In response to ECS, this gives rise to a phase III oscillation that does not have the ictal spikes characteristic of epileptic seizures. Ictal spikes in epileptic seizures are thought to reflect cortical responses to the oscillating thalamic drive (Meeren et al., 2002; Polack et al., 2009). Interestingly, a combination of CSD induced plasticity and the phase III oscillation (Figure 3D) could result in a selective strengthening of thalamocortical synapses. It has been speculated that such an increase is the mechanism that underlies the therapeutic efficacy of ECT (Wei et al., 2020). Assuming psychosis is the result of thalamocortical coupling that is too weak in frontal areas of the cortex (Keller and Sterzer, 2024), this mechanism of action could explain both the efficacy of ECT in treating psychosis and why patients with psychotic depression have more favorable ECT treatment outcomes than those with nonpsychotic depression (Petrides et al., 2001).

Future directions

Unresolved is the question of how to reconcile the non-specific nature of CSD with the cell type specific effects that characterize rTMS (Gongwer et al., 2025), antidepressant (Shao et al., 2025; Vesuna et al., 2020) and antipsychotic drugs (Heindorf and Keller, 2024; Lupori et al., 2026). Several lines of evidence suggest that the effects of CSD may also be cell type specific. First, CSD is easier to trigger in superficial than in deeper layers of cortex (Leo and Morison, 1945), and affects superficial somata for a significantly longer duration (Figure 4). This aligns with the evidence that schizophrenia is associated to a number of plasticity-related genes whose expression is selectively enriched in upper cortical layers (Batiuk et al., 2022). Second, the large spectrum of time courses of calcium responses to CSD (Figure 4) could be explained by cell type intrinsic mechanisms, such as differential NMDA receptor expression (Buchanan et al., 2012). Testing this hypothesis will require cell type specific chronic calcium imaging following ECS.

The immediate clinical challenge stemming from our findings will be to determine whether the role of CSD is therapeutically beneficial or potentially deleterious. Optimizing ECT involves balancing the likelihood of positive therapeutic outcomes while minimizing adverse cognitive effects, most notably retrograde amnesia and impairments in autobiographic memory (Porter et al., 2020; Sackeim et al., 2007). For example, there is an optimal window for the stimulation charge: while a minimum stimulation charge is required for therapeutic benefit, a stimulation charge significantly exceeding the phase III oscillation threshold is linked to increased cognitive side effects (Delva et al., 2000; Sackeim et al., 1987). This could be explained if a higher stimulation charge is more likely to trigger CSDs in brain regions more distant from the stimulation electrodes. For example, it is conceivable that a CSD in frontal cortex is therapeutically beneficial while a CSD in hippocampus is responsible for retrograde amnesia. In mice, the ECS driven CSD likely invades hippocampus (Figure 5) (Mitlasóczki et al., 2025). If so, this might open up potential for pharmacological interventions to prevent a CSD invasion of hippocampus during ECT, e.g. with NMDA receptor antagonists that preferentially affect hippocampus (Mutel et al., 1998). By restricting CSD to the neocortex, it may be possible to prevent retrograde amnesia without compromising the cortical plasticity effects required for therapeutic outcomes.

Relevance

Our findings suggest that CSD, a currently unmonitored neuronal response to ECS, appears to be the primary driver of ECT-induced neuroplasticity. Ultimately, we propose that monitoring and facilitating CSD during ECT may unlock the possibility to transform psychiatry’s most potent intervention into a targeted and reliable therapy with minimal side effects.

Methods

Mice

A total of 66 mice (27 male, 39 female), 7 – 22 weeks old at the start of the experiment, were used. Most mice were C57BL/6, but for two-photon prism imaging, the cohort also included PlexinD1-CreERT2 heterozygous mice. In the PlexinD1-CreERT2 mice, tamoxifen food was administered to induce ChrimsonR expression in the retrosplenial cortex for a separate experiment. See Table S2 for details of mouse inclusion for the different experiments. Mice were group-housed in a vivarium (light/dark cycle: 12/12 hours), and experiments carried out in light cycle. In female mice, we did not time experiments to a specific phase of the estrous cycle. All animal procedures were approved by and carried out in accordance with the guidelines of the Veterinary Department of the Canton of Basel-Stadt, Switzerland.

Surgery

For all surgical procedures, mice were anesthetized with a mixture of fentanyl (0.05 mg/kg; Actavis), midazolam (5.0 mg/kg; Dormicum, Roche), and medetomidine (0.5 mg/kg; Domitor, Orion) injected intraperitoneally. Analgesics were applied perioperatively (2% lidocaine gel, meloxicam 5 mg/kg) and postoperatively (3.25 mg/kg; Ethiqa XR). Eyes were covered with ophthalmic gel (Virbac Schweiz AG). For widefield imaging, we implanted crystal skull (Kim et al., 2016) or thinned skull (Yang et al., 2010) cranial windows. Briefly, the perimeter of the skull plate overlying the cortex was thinned using a dental drill until it could be removed with forceps (crystal skull) or only thinned in its entirety until transparent but not removed (thinned skull). To protect the brain, a glass or acrylic polymer window was then implanted. For two-photon imaging, a cranial window was implanted over primary visual cortex (V1) as previously described (Keller et al., 2012; Leinweber et al., 2014). Briefly, a 4 mm craniotomy was made over the right V1, centered 2.5 mm lateral and 0.5 mm anterior to lambda. The exposed cortex was sealed with a 3 mm or 4 mm circular glass coverslip. For prism imaging, a 0.7 mm glass prism (Tower Optical Corporation) was attached to the coverslip with UV-curable glue (Norland Optical Adhesive), and a small incision of cortical tissue was made to allow for the insertion of the prism into the cortex. Cranial windows were sealed using gel superglue (Ultra Gel, Pattex), and, for widefield windows, cyanoacrylate adhesive (Vetbond, 3M) (Kauvar et al., 2020). The remaining exposed surface of the skull was covered with cyanoacrylate adhesive, and a titanium head bar was fixed to the skull using dental cement (Paladur, Heraeus Kulzer). After surgery, anesthesia was antagonized by a mixture of flumazenil (0.5 mg/kg; Anexate, Roche) and atipamezole (2.5 mg/kg; Antisedan, Orion Pharma) injected intraperitoneally.

Virus injections

For widefield experiments, we deposited an AAV vector with PHP.eB capsid (Chan et al., 2017) retro-orbitally (4 μl per eye of at least 1014 GC/ml) to drive expression throughout cortex of GCaMP6s (AAV-PHP.eB-Ef1α-GCaMP6s-WPRE) or EGFP (AAV-PHP.eB-Ef1α-EGFP-WPRE). For two-photon experiments, the AAV was injected locally to drive expression of GCaMP6f (AAV2/9-Ef1α-GCaMP6f-WPRE). Imaging commenced at the earliest 3 weeks after AAV injection.

Electroconvulsive stimulation (ECS) in mice

For all ECS experiments mice were anesthetized with isoflurane (2%) using 95% O2 as a gas carrier (1 L/min). For myorelaxation we injected succinylcholine chloride (1 mg/kg; Sigma-Aldrich) intraperitoneally. Mice were head-fixed, and ECS was delivered through auricular electrodes (Ugo Basile model #57802) using a pulse generator (Ugo Basile model #57800-001). A conductive salt-free electrolyte gel (Spectra 360 gel, Parker Laboratories) was applied to the electrodes to improve conductivity. Stimulation consisted of a 1 s long pulse train with a frequency of between 25 and 100 Hz, and a current amplitude of between 15 and 75 mA. Pulse width was 0.5 ms, which resulted in a total charge range of between 0.187 and 3.750 mC. These parameters reliably evoke tonic-clonic convulsions in mice (Hagen et al., 2015; van Buel et al., 2017). For comparison, patient ECT parameters used here were typically as follows: train duration up to 8 seconds, pulse width 0.5-1 ms, a fixed current amplitude of 900 mA, pulse frequency between 40-60 Hz, charge maximum 1000 mC. Thus, stimulus duration, current amplitude and resulting total charge are adapted to the mouse, but frequency/pulse-width are within clinical range (Peterchev et al., 2010). Of note, the stimulation polarity was constant for mice ECS, whereas bidirectional pulse train (alternating polarity) was used for patient ECT. For sham ECS experiments, we used an identical procedure with no current delivery. Up to 6 ECS sessions were administered at intervals of 48 hours. Stimulation intensities and pulse characteristics were randomized across subjects to eliminate order effects.

Electroconvulsive therapy (ECT) in humans

The retrospective treatment data of 28 random patients (10 male, 18 female, all > 18 years old), who underwent ECT at the Zentrum für Affektive, Stress-und Schlafstörungen (ZASS), University Psychiatric Clinics (UPK), Basel due to a psychiatric indication were included. Patients signed a general consent form (approved by the local ethics committee). All patient procedures were approved by and carried out in accordance with the Ethikkommission Nordwest-und Zentralschweiz (EKNZ). One randomly chosen treatment within the first half of the ECT acute treatment series (i.e. < 7th treatment) per person was considered for analysis. Restimulation treatment data were excluded. Medical indication for ECT, patient information, treatment preparation and anesthesia followed standardized clinical protocols according to current standards of care. ECT was conducted with bifrontal stimulation using the Thymatron® System IV device (Somatics, LLC. Lake Bluff, IL, USA) and disposable adherent stimulus electrodes (EPAD, Thymapad®), as previously described (The Practice of Electroconvulsive Therapy Third Edition, 2025), following Kellner’s ECT guidelines (Kellner, 2018). Briefly, patients were anesthetized with either propofol or etomidate (Braun Medical AG) and myorelaxed with succinylcholine chloride (Amino AG). Skin surface was cleansed using an alcohol-containing solvent and abrasive gel. A conductive and adhesive liquid (Signaspray®, Parker Laboratories) was then applied to the skin. EEG was recorded at 200 Hz using bilateral frontomastoid placement and filtered using a band-pass with 3dB points at 2 and 25 Hz. We excluded the first two seconds of recording following the stimulation to remove a capacitive coupling artifact. ECT was performed about 4 min after the injection of etomidate and 5 min after the injection of propofol, using the ‘double dose’ stimulation protocol. ECT sessions included in the analysis were performed by multiple clinicians with standardized preparation and stimulation procedures. Data were measured using the Thymatron during each session, including stimulation parameters and EEG derived seizure quality parameters. In addition, we had access to formalized clinical documentation listing the administered medication and manually assessed seizure duration. Digitalization of stored ECT data for the current analysis was performed after December 2020 using GENET-GPD (Freundlieb et al., 2023). The digitalized treatment data were derived from the GENET-GPD database. Patient data were stored in the UPK patient case files in compliance with accepted data protection and record-keeping regulations.

Electroencephalography (EEG) in mice

Mice were anesthetized as described above, and the glass or polymer window was removed. An EEG multielectrode array with 30 recording sites (EEG mouse-30-A-10-H32, Neuronexus) was used to record neural activity across the surface of the dorsal cortex. Signals were amplified using a 32 channel headstage (RHD 32ch, Intan Technologies) and digitized at 30 kHz using an Open Ephys acquisition board (Siegle et al., 2017). During ECS delivery, the EEG array was physically disconnected by an electronically triggered relay to prevent electrical arcing between electrodes. EEG data was downsampled to 1 kHz and band-pass filtered between 0.5 and 10 Hz. We excluded the first two seconds of recording following EEG reconnection to remove a capacitive coupling artifact (Figure S1). As the concurrent widefield calcium signal provided an unambiguous polarity reference, the polarity of the EEG signal is such that positive voltage deflections corresponded to increases in fluorescence.

Widefield imaging

Widefield imaging experiments were conducted on a custom-built macroscope with objectives mounted face-to-face (Nikon 85 mm/f1.8 sample side, Nikon 50 mm/f1.4 sensor side). A 470 nm LED (Thorlabs) powered by a custom-built LED driver was used to excite fluorescent indicators through an excitation filter (SP490, Thorlabs) reflected off of a dichroic mirror (LP490, Thorlabs) placed parfocal to the objectives. The fluorescence was collected through a 525/50 emission filter on a sCMOS camera (PCO edge 4.2). LED illumination was adjusted with a collimator (Thorlabs SM2F32-A) to homogenously illuminate cortical surface through the cranial window. Images were acquired at 100 Hz frame rate. Raw images were cropped on the sensor and data was stored to disk with custom-written LabVIEW (National Instruments) software, resulting in an effective pixel size of 60 μm2 at a resolution of 1108 by 1220 pixel (1.35 MP).

Two-photon imaging

Two-photon calcium imaging was performed using custom-built microscopes (Leinweber et al. 2014). The illumination source was a tunable femtosecond laser (Insight, Spectra Physics or Chameleon, Coherent) tuned to 930 nm. Emission light was band-pass filtered using a 525/50 filter and detected using a GaAsP photomultiplier (H7422, Hamamatsu). Photomultiplier signals were amplified (DHPCA-100, Femto), digitized (NI5772, National Instruments) at 800 MHz, and band-pass filtered at 80 MHz using a digital Fourier-transform filter implemented in custom-written software on an FPGA (NI5772, National Instruments). The scanning system of the microscopes was based on a 12 kHz resonant scanner (Cambridge Technology). Images were acquired at a resolution of 750 by 400 pixels (60 Hz frame rate) with a 16x, 0.8 NA water immersion objective (Nikon). We imaged layer 2/3 at a depth of 100-250 μm with a field of view of 375 by 300 µm. For prism imaging, to enable imaging of as much cortical depth as possible, the field of view was expanded to 270 by 960 µm, at a resolution of 750 by 800 pixels. This resulted in an effective frame rate of 30 Hz. Laminar boundaries were estimated from the density of GCaMP6-expressing cells, which typically peaks in layer 2/3 and upper layer 5 (Andermann et al., 2013).

Image processing

For analysis of chronic widefield macroscope imaging, raw movie data was manually registered across days by aligning subsequent mean projections of the data to the first recorded image sequence. In the raw 1108 by 1220 pixel images, we tiled the cortex with 818 ROIs each with a fixed location relative to bregma and lambda. This resulted in an average ROI size of 25 by 25 pixels (approximately 230 by 230 μm). Each ROI was also registered against the Allen Institute Mouse Brain Coordinate Framework (Wang et al., 2020). In concurrent EEG and widefield recordings, we defined annular ROIs centered on each of the 30 electrode sites. The inner boundary corresponded to the electrode footprint with a 50 μm buffer (approximately 60 pixels in diameter; approximately 550 μm) and was excluded from analysis, while the outer boundary defined a circular region with a diameter of approximately 80 pixels. Two-photon calcium imaging data were processed as previously described (Attinger et al., 2017; Keller et al., 2012). Briefly, raw images were full-frame registered to correct for lateral brain motion. Somata ROIs were manually selected based on mean and maximum fluorescence images. For each somata ROI, a corresponding neuropil ROI of equal pixel area was created as a surrounding annulus, maintaining a 3-pixel exclusion buffer and avoiding overlap with other defined ROIs. No neuropil subtraction to somata activity was done. Activity was calculated as the ΔF/F0, where F0 was the median fluorescence of the recording before ECS (typically 30 seconds). ECS recordings lasted for 5 min.

Histology

Mice underwent a single ECS or sham procedure while being monitored for calcium events through widefield imaging and were returned to their home cage for one hour. A subset (11/17) of mice was instead kept head-fixed and imaged for 1 hour to verify that no additional calcium events occurred. Mice were then anesthetized with a mixture of fentanyl (0.05 mg/kg; Actavis), midazolam (5.0 mg/kg; Dormicum, Roche), and medetomidine (0.5 mg/kg; Domitor, Orion), and transcardially perfused with phosphate-buffered saline (PBS) solution and subsequently with ice-cold 4% paraformaldehyde (PFA) in PBS. Brains were extracted and post-fixed overnight in 4% PFA at 4 °C. They were then embedded in 4% agarose and coronally sectioned through the entire anteroposterior extent of the cortex at (50 µm thickness) using a vibratome (Leica VT100S). We collected every fourth section for further processing, resulting in an effective sampling step of 200 µm. Free-floating sections were then processed for Fos immunohistochemistry. Briefly, sections were blocked for 2 h at room temperature in a solution containing 5% donkey serum (BSA, Sigma-Aldrich) and 0.5% Triton X-100 in PBS, then incubated overnight at 4 °C with a solution containing anti-Fos antibody (226008, Synaptic Systems, 1:1000), 1% Donkey Serum and 0.1% Triton X-100 in PBS. Wells were then rinsed 3 times in PBS (10 min each) and incubated with a solution containing donkey polyclonal anti-rabbit antibody (A31573, Invitrogen, Alexa Fluor™ 647 conjugate, 1:300), 1% Donkey Serum and 0.1% Triton X-100 in PBS for 2 h at room temperature. Following additional 10-minutes PBS washes, nuclei were counterstained with DAPI (Sigma-Aldrich, 1:1000) for 10 minutes. Finally, sections were mounted on microscopy slides with a mounting medium (VECTASHIELD, H-100, Vector Laboratories), sealed with nail polish, and stored at 4°C. Images were then acquired with an Axioscan Z1 (Zeiss) microscope using an apochromatic 10x/0.5 M27 objective (Zeiss). Fluorophores were excited with 422 nm and 668 nm X-Cite Xylis LED. The emission wavelength were filtered with a BP 445/50 filter for DAPI and a BP690/50 filter for Alexa 647. Images were collected with an Hamamatsu ORCA flash 4.0V camera at a pixel size of 0.650 µm.

Tissue registration and quantification

Images were aligned to the Allen Mouse Brain Common Coordinate Framework (CCFv3-2017) as described previously (Lupori et al., 2023). All preprocessing and registration steps, unless stated otherwise, were carried out using custom-written MATLAB (MathWorks) code. For each brain, images were ordered along the posterior-to-anterior direction. Based on a reference mark identifying the right side and/or tissue irregularities, images were mirrored vertically so that all matching hemispheres would be on the same side. Binary masks to restrict quantification to brain tissue were automatically generated from the DAPI images by training a pixel classifier in Ilastik (Berg et al. 2019). Each mask was then visually inspected and manually adjusted based on the Fos images, to remove regions containing tissue damage or staining artifacts. Left-right binary masks defining which hemisphere a pixel belongs to were then generated by manually drawing a segment at the midline of each slice. Using the DAPI images, each slice was matched to a plane in the CCFv3 though one first step of global alignment using QuickNII v2.2 (Puchades et al. 2019) and a second step of local non-rigid transformation with VisuAlign v0.9 (RRID:SCR_017978, VisuAlign). For cell counting, Fos channel images were first pre-processed with a rolling ball filter (50 pixels or 32.5 μm radius) for background removal. Cell detection and counting were then performed on the filtered images with custom scripting inside QuPath v.0.6.0 (Bankhead et al. 2017). Briefly, we used the inbuilt automatic cell detection tool to identify Fos-positive cells using the median fluorescence of the slide as a threshold. A random forest classifier was then trained to remove detection artifacts. For each cell, we calculated mean pixel intensity and x-y coordinates of the centroid. For all regions descending from isocortex, as per the hierarchy defined in the CCFv3, we computed cell density (number of cells/total area in mm2), mean cell intensity, and energy (cell density*mean cell intensity) (Lein et al., 2007; Lupori et al., 2023). For each metric, the value was then normalized by the average of that metric across all sham-treated mice.

Data analysis

All data analysis was performed with custom-written MATLAB (MathWorks) code.

When visualizing data with a box plot, the central box indicates the interquartile range (IQR), spanning from the first (Q1) to the third (Q3) quartiles. Whiskers extend to data points not considered outliers. Outliers were defined as data points lying outside of the Q1 – 1.5 IQR to Q3 + 1.5 IQR range. A central horizontal line indicates the median and an open circle the mean of the distribution. To represent the underlying probability density, a Kernel Density Estimate was computed using a Parzen-window method with a normally distributed kernel.

In Figure 1 and Figure 6, spectrograms were computed using a 2 s sliding window with 95 % overlap of the band-passed signal between 0.5 and 10 Hz. Power spectral density was estimated for each channel or ROI independently and converted to decibel scale. Oscillations were detected in the 1-5 Hz band by computing the Hilbert envelope of the EEG signal. Oscillation onsets were defined as the first crossing of the 90th percentile of the envelope after ECS. Oscillation terminations were defined as the first time the envelope dropped below a baseline threshold. Patient EEG is interrupted rapidly after the termination of oscillations. To account for the differences in recording length, the termination threshold was set as the 50th percentile Hilbert envelope for mice and as the 5th percentile for patients. For patient EEG, this detection method yielded oscillation durations not statistically different to those computed by the proprietary Thymatron algorithm (hierarchical bootstrap with resampling, p = 0.268). Power-weighted mean frequencies were computed as the sum of power values in a frequency, multiplied by that frequency. We reported the median values in a 2 s window centered on the global frequency peak and in the final 2 s of the detected oscillation. For mouse widefield data, the onsets and offsets time were computed from the concurrent EEG signal. We excluded frequencies between 7 and 9 Hz, if they exceeded 90 μV; these regions were dilated by 2 s and masked-out from further analysis. We completely excluded EEG electrodes and their associated widefield ROI if the onset/offset values were non-finite or if the total duration of oscillation was less than 1s, resulting in the exclusion of 78/480 (16%) of electrodes in our dataset.

In Figure 2 and Figure S3, calcium event speed was estimated through the local fluorescence gradient. Peak latencies were defined as the global maximum occurring 10 to 200 seconds after local wavefront onset. For each ROI, raw fluorescence values were low-passed filtered below 0.5 Hz and a plane (t = Sxx + Syy + t0) was fitted to the peak latencies of neighbors within a 0.8 mm radius using bisquare weighting. This determined a local slowness vector, the magnitude of which was inverted to determine speed. Speed values deviating more than 3.5 median absolute deviation from the session median were excluded. For two-photon imaging, propagation speed was calculated as the median velocity of calcium events across pairs of somata ROIs separated by at least 200 µm.

In Figure 2 and Figure S4, the period of direct ECS stimulation is defined as the first 5 s post ECS onset, based on the duration reported in Figure S1.

In Figure 2, 3 and Figure S3, the onset of a calcium event was defined as the first crossing of 150% the mean calcium activity before ECS. Only events occurring 10 s after ECS were considered as calcium events. For analysis of time to peak, we excluded ROIs for which there was no clear separation between direct ECS response and calcium event. To determine the presence of an oscillation, we averaged the calcium traces across hemispheres and used the same algorithm used in Figure 1 and Figure 6. To characterize the probability of calcium event or oscillation as a function of ECS parameters, we fitted a linear slope with ordinary least squares method. 95% confidence bands for the mean predicted response were computed across the range of predictor variables.

In Figure 2, to generate isochrones, we set a ROI-wise threshold to 150% of the maximum ECS baseline activity. The first time at which each ROI crossed this threshold was used to construct a time-of-activation map, which was then filtered with a normal distribution. Contour lines were extracted at 2 second intervals from the onset of calcium event until 50 s post-onset.

In Figure 4, cortical layers were separated in supragranular and infragranular, based on empirical distribution of cell densities, which shows a clear gap in cortical layer 4. To compare the activity of somata with their neuropil, we aligned each somata with the peak of its neuropil during calcium events. This alignment ensures that the propagation delay across the field of view is not taken into account. To compare supragranular and infragranular layers, somata were instead aligned to the max of a neuropil trace, averaged across all neuropil ROIs in the entire recording. To quantify the dynamics of GCaMP6f fluorescence following the calcium event, we computed a linear recovery slope with least-square ordinary fit in the 2 to 60s seconds following peak somata signal.

In Figure 6, the mid-ictal amplitude was computed as the mean of all oscillation peaks in a 4 s window centered at the duration of the oscillation (Krystal et al., 1995). The postictal suppression index was defined as the ratio of the median 1-30 Hz envelope amplitude post-oscillation over intra-oscillations. Welch’s periodogram method was used to compute the inter-hemispheric coherence between pairs of symmetric electrodes. We reported the values of coherence in the 1 to 5 Hz band.

In Figure S2, to illustrate the synchronicity of EEG and calcium signals, oscillation peaks in the EEG channel were detected using the Automatic Multiscale Peak Detection algorithm, which is suited to non-stationary signals, such as EEG, without requiring manual parameter tuning or pre-filtering (Scholkmann et al., 2012).

Statistical analysis

All statistical information for the tests performed in the manuscript is provided in Table S1. Analyses accounted for the nested structure of the data consisting of recording units (neurons, EEG channels or ROIs) within subjects (mice or patients) using hierarchical bootstrap (Saravanan et al., 2020). Briefly, we first resampled with replacement data at the subject level and then resampled with replacement a second time at the recording unit level. We then computed the mean of this bootstrap sample and repeated this N times to generate a bootstrap distribution of the mean estimate. For all statistical testing the number of bootstrap samples (N) was 10 000, for plotting bootstrap mean and standard error response curves it was 1000. The bootstrap standard error is the 68% confidence interval (1 SD, 15.8th percentile to 84.2nd percentile) in the bootstrap distribution of the mean.

To account for the false discovery rate in multiple comparisons, we applied an adjustment to p-values (Benjamini and Hochberg, 1995). Briefly, p-values were sorted in ascending order and multiplied by the ratio of the total number of comparisons within a data group (for example, all bootstrap tests for a specific EEG metric) to their sorting rank. The final significance cutoff was determined by finding the largest p-value smaller than its rank-specific threshold, at which point all p-values with a higher rank (i.e., those with smaller p-values) were also rejected.

For linear regression analysis, we computed the F-statistic to test the significance of the model. This test evaluates whether the slope is significantly different from zero by comparing the model degrees of freedom (the number of predictors, here, one) against the degrees of freedom for error (the number of observations minus the parameters for the intercept and slope). The resulting F-statistic represents the ratio of explained variance to unexplained variance, quantifying the portion of the data’s variability accounted for by the linear regression. For multivariate linear regression analysis, we measured the probability of occurrence of a calcium event as a function of stimulation amplitude, frequency, and their two-way interaction. For each of those three parameters, we performed a two-sided t-test against the null hypothesis that this parameter is equal to zero.

eLife Assessment

This study addresses a recent discovery by others that electroconvulsive therapy (ECT) generates seizure activity and spreading depolarization (SD), reflected in large increases in calcium, which can be followed by imaging calcium fluctuations in neurons. This work is useful. However, the evidence to show that SD, rather than seizures, confers the neuroplastic and other therapeutic effects of ECT is incomplete.

Reviewer #1 (Public review):

The work corroborates the idea, recently suggested by Rosenthal et al. (2025), that spreading depolarization is involved in the mechanisms of electroconvulsive therapy. Using a mouse model of electroconvulsive therapy and various sophisticated approaches to visualize cortical activity, the authors provide an extensive description of traveling calcium waves induced by electroconvulsive stimulation. The study confirms that the calcium events have properties typical of cortical spreading depolarization and seeks to show that the calcium/SD waves mediate therapeutic and neuroplastic effects of electroconvulsive therapy. The authors find that after electroconvulsive stimulation associated with calcium/SD waves, Fos expression increases widely; in the cortex, this increase is localized to the hemisphere affected by calcium waves. They show that some EEG predictors of the beneficial effects of electroconvulsive therapy correlate with the occurrence of calcium/SD waves. Despite the solid methodology and the study's interesting, its conclusions are not fully supported by the data.

In particular:

(1)The title of the paper claims that "electroconvulsive stimulation drives cortical spreading depolarization dependent immediate early gene expression". However, immunohistochemical staining shows that Fos expression increases not only in the cortex but also in many subcortical regions, including the hippocampus and amygdala (Figure 5A). Really, conventional electroconvulsive therapy stimulates nearly the entire brain volume and induces generalized seizure activity that can trigger SD not only in the cortex but also in other brain sites. Therefore, regions beyond the cortex can also drive the effects of electroconvulsive therapy. Next, the authors use Fos staining as a marker of neuronal plasticity. However, Fos is also a marker of preceding neuronal activation. As electroconvulsive stimulation, seizures, and SD are associated with high neural activity, it is unclear whether the observed Fos upregulation results from the prior activation or heralds the subsequent plastic changes. Other markers of neuroplasticity (e.g., BDNF) should also be examined.

(2) Postictal EEG suppression is one of the most promising correlates of positive clinical outcomes after electroconvulsive therapy. Cortical SD is also tightly coupled with suppression of neuronal activity in affected regions. Although the authors report that postictal suppression is stronger after stimulations with cortical SDs than without SDs, the cortices affected (ipsi) and unaffected (contra) by unilateral cortical calcium/SD events exhibit identical suppression (Figure 6F). The result contradicts established knowledge in the field. If the calcium events are cortical SDs, they should induce EEG suppression only in the affected hemisphere.

(3) The study states a beneficial role of calcium/SD waves in ECS effects. However, SD alters numerous aspects of brain function, leading to a range of effects that can underlie side effects as well. Assessment of the behavioral effects of stimulation with and without calcium/SD waves can help clarify the issue.

The results of the work suggest that cortical SD can contribute to electroconvulsive therapy-related mechanisms and help to optimize the stimulation parameters to achieve maximal therapeutic effect.

Reviewer #2 (Public review):

Summary:

This manuscript addresses the question of mechanisms underlying the therapeutic effects of electroconvulsive therapy (ECT). Clinical efficacy of ECT in major depression (and other disorders) is well established and has often been assumed to be a direct consequence of seizure activity generated by the current application. However, as the authors point out, this explanation is unsatisfactory. A recent study (Rosenthal et al., 2025) provided evidence that ECT generates a wave of cortical spreading depolarization (CSD) in mice, and initial evidence that similar events were generated in patients undergoing ECT. Based on their observations, Rosenthal et al. proposed that CSD, rather than seizure, may engage plasticity mechanisms that contribute to the brain's clinical response to ECT. The current study adds to that prior work by reporting other consequences of CSD, in addition to sustained Ca2+ elevations. The current study also links EEG characteristics immediately following the ECT with the likelihood of generating a CSD, which can help optimize ECT parameters.

Strengths:

An important research topic, linking a large set of rodent studies with a limited clinical EEG data set.

The data acquisition and analyses appear to be of very high quality, and the main results are well illustrated.

Association between EEG characteristics linked to good clinical outcome matched by mouse EEG data linked to CSD.

Characterization of multiple consequences of CSD following ECT in the mouse brain.

Weaknesses:

The main characterization of CSD propagation comes from GCaMP Ca2+ measurements, as previously reported (Rosenthal et al., 2025). That prior study also provided key electrophysiological evidence of CSD with a DC shift after ECT in mice (supplemental data). Given the prior evidence for ECT-CSD, the additional measures shown in the current manuscript are fully expected. Thus, the 2-photon imaging of Ca2+ elevations following CSD (Figure 4) is consistent with prior 2-photon imaging studies of CSD, and the complex hemodynamic and pH changes are expected to contribute to propagation of EGFP fluorescence changes (Supplemental Figure 5). These data are well presented, but, contrary to the results section here, these results appear confirmatory rather than necessary to build a case that the key event generated by ECT is a CSD.

The authors state that "our conclusion that CSD is the primary driver of plasticity is based on its role in driving Fos expression" (line 472). Related to the point above, there is already a very well-established literature showing that CSD leads to rapid and robust Fos expression in rodent cortex, so this is fully consistent with prior work. The prior work, CSD-fos work, should be summarized and/or cited more clearly in the manuscript. Showing that Fos increases only in the hemisphere where there is a large CSD-Ca2+ wave is a clear demonstration of this. While Fos increases can certainly be well linked to plasticity in some experimental paradigms, the implication that Fos increases underlie CSD-induced plasticity and possibly therapeutic effects of ECT is not appropriate. Fos increases after CSD are a reliable marker of the very strong neuronal activation that occurs, but Fos increases are not specific for plasticity and can be activated by challenges that do not generate synaptic plasticity. A range of other gene expression changes have been identified with CSD and may contribute to adaptive plasticity; these could be mentioned alongside speculation about Fos. To support the main conclusions of this paper about CSD driving plasticity via Fos, Fos knockout or knockdown studies are needed, as has been used in prior plasticity studies.

Reviewer #3 (Public review):

Summary:

This manuscript combines widefield calcium imaging, electroencephalography, 2-photon imaging, and immunohistochemistry in mice to re-demonstrate that electroconvulsive stimulation (ECS) induces a seizure followed by cortical spreading depolarization, as previously shown. The putative novel finding - which is not unexpected - is that ECS is also correlated with increased expression of the immediate early gene cFOS, although this has also been shown previously. The authors speculate that CSD drives cFOS expression, which might contribute to the therapeutic effects of ECT; however, experiments performed do not provide causal evidence for this hypothesis. Instead, the authors use expression of cFOS - a nonspecific activity-dependent gene induced in various pathological and non-therapeutic contexts - as a proxy for plasticity and/or therapeutic effect. Hence, overall, the significance of the findings is limited and primarily serves to replicate prior work, with the evidence evaluated as incomplete.

Strengths:

The experiments are generally well executed from a technical perspective.

Main Weaknesses to be addressed in revision:

(1) The main findings of this paper are replication experiments of prior work, and thus, the novelty and significance of this manuscript are relatively limited.

- It is already known that the mean frequency of ECT-induced seizures decays between peak and offset in humans (Stuiver et al. Clin Neurophysiol. 2026 Jan:181:2111439. doi: 10.1016/j.clinph.2025.2111439) and mice (Murakami et al. J Pharmacol Sci 2008 Jan;106(1):78-83. 10.1254/jphs.FP0071453), which the authors re-demonstrate in Figure 1.

- It has already been demonstrated that ECT in mouse models induces lateralized CSD waves in a manner that depends on stimulation parameters and the initial evoked response during stimulation (Rosenthal et al. Nat Comm. 2025 May 18;16(1):4619. doi: 10.1038/s41467-025-59900-1); the authors replicate this in Figures 1, 2, 3, 6.

- It is already widely established that EEG and calcium signals are highly concordant in mouse brain physiology, as shown in Figure 1. It is already known that CSD propagates from supragranular to granular and infragranular layers (Zakharov et al. Epilepsia. 2019 Dec;60(12):2386-2397. doi: 10.1111/epi.16390) as shown in Figure 4.

- It is already known that CSD waves induce cFOS expression (e.g., Dell'Orco et al. Front Cell Neurosci. 2023 Dec 14:17:1292661. doi: 10.3389/fncel.2023.1292661; Hermann and Hossman. Neuroscience. 1999 Jan;88(2):599-608. doi: 10.1016/s0306-4522(98)00249-8) as the authors replicate in Figure 5.

Minimally, the authors should revise claims regarding novelty, as the manuscript, as written, is misleading to a reader not familiar with the field. There is limited innovation in re-demonstrating that these events are seizures and that they involve spreading depolarization.

(2) The authors frame their hypothesis that CSD could be a potential mediator of the therapeutic effects of ECT, but they do not measure therapeutic effects or directly test this hypothesis. The principal advancement of the paper is showing that ECT-induced CSD triggers hemisphere-specific cFOS expression as a proxy of plasticity. However, it is already known that CSD induces cFOS expression (as noted above). The observation that cFOS expression was induced only by CSD, not by the initial seizure, is likely a byproduct of the greater activity induced by CSD than by seizure. cFOS expression is nonspecific to plasticity or therapeutic effects and can be triggered by many non-therapeutic interventions. The cFOS data thus do not meaningfully measure therapeutic plasticity. The authors also selectively cite references suggesting that EEG metrics such as seizure duration predict positive therapeutic outcomes, but this link is controversial and not well established in the clinical literature.

Minor Weaknesses:

(3) For the n=3 mice used for concurrent 2P imaging with microprism implant, these animals also had ChrimsonR co-expression, but there are no optogenetic studies described in this paper, which is confusing. Yet, this co-expression introduces a significant confound, as GCaMP6 emission (525/50nm band in this study) will overlap substantially with the ChrimsonR excitation spectrum. Thus, the fluorescence emission used to image these neurons may be optogenetically activating them at the same time. Please explain.

(4) Incision of the cortex for implantation of a prism is a significant cortical injury that likely induces CSD instantaneously and may change the propensity for CSD in subsequent recordings. Please comment on this limitation and address how much time elapsed after surgery before imaging.

(5) Method details are missing or insufficiently described for location, titer, and injection strategy for 2-photon experiments.

(6) Given the wide range of parameters used for ECS in mice and ECT in humans, the authors should provide tables for what stimulation parameters were used for each recording. These protocols were chosen manually rather than randomly or systematically, which introduces confounding factors into analyses that use parameters as an independent variable.

(7) While much of the cFOS staining after unilateral CSD shows hemisphere-specific asymmetry, several regions (piriform cortex, amygdala, thalamus) do appear to have bilateral cFOS expression. Please comment on this.

(8) The discussion states: "If CSD accounts for plasticity effects, triggering a CSD in a non-seizure context may be sufficient to elicit therapeutic effects. This is supported by the clinical success of ultra-brief stimulation treatments that do not cause seizures, such as rTMS with accelerated protocols, which achieves treatment efficacy on par with ECT for major depressive disorder". Are the authors implying that TMS induces CSD? What evidence supports this idea?

(9) This statement - "Assuming psychosis is the result of thalamocortical coupling that is too weak in frontal areas of the cortex" (lines 583-585) - may be overly speculative.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

The work corroborates the idea, recently suggested by Rosenthal et al. (2025), that spreading depolarization is involved in the mechanisms of electroconvulsive therapy. Using a mouse model of electroconvulsive therapy and various sophisticated approaches to visualize cortical activity, the authors provide an extensive description of traveling calcium waves induced by electroconvulsive stimulation. The study confirms that the calcium events have properties typical of cortical spreading depolarization and seeks to show that the calcium/SD waves mediate therapeutic and neuroplastic effects of electroconvulsive therapy. The authors find that after electroconvulsive stimulation associated with calcium/SD waves, Fos expression increases widely; in the cortex, this increase is localized to the hemisphere affected by calcium waves. They show that some EEG predictors of the beneficial effects of electroconvulsive therapy correlate with the occurrence of calcium/SD waves. Despite the solid methodology and the study's interesting, its conclusions are not fully supported by the data.

In particular:

(1) The title of the paper claims that "electroconvulsive stimulation drives cortical spreading depolarization dependent immediate early gene expression". However, immunohistochemical staining shows that Fos expression increases not only in the cortex but also in many subcortical regions, including the hippocampus and amygdala (Figure 5A). Really, conventional electroconvulsive therapy stimulates nearly the entire brain volume and induces generalized seizure activity that can trigger SD not only in the cortex but also in other brain sites. Therefore, regions beyond the cortex can also drive the effects of electroconvulsive therapy.

This is correct, our claim is that ECS drives spreading depression dependent IEG expression in cortex. We are not claiming this is the only consequence of ECS. The extracortical response to ECS could contribute to the treatment, and we discuss this in the hippocampus specifically (see Discussion). In current clinical practice, cortical surface EEG is used as a biomarker for predicting treatment outcome. Given that the CSD is necessary to drive Fos expression in cortex, we think it is warranted to speculate that monitoring CSD during ECT is worth exploring as a potentially superior biomarker of treatment outcome. Especially given that this is readily feasible (Discussion).

Next, the authors use Fos staining as a marker of neuronal plasticity. However, Fos is also a marker of preceding neuronal activation. As electroconvulsive stimulation, seizures, and SD are associated with high neural activity, it is unclear whether the observed Fos upregulation results from the prior activation or heralds the subsequent plastic changes. Other markers of neuroplasticity (e.g., BDNF) should also be examined.

The idea that Fos is a marker of neuronal activity is outdated. The primary correlate of Fos expression is neuronal plasticity and learning-related circuit modifications, rather than just neuronal activity – see e.g. (Bolhuis et al., 2001; de Hoz et al., 2018; Fleischmann et al., 2003; Kimpo and Doupe, 1997; Mahringer et al., 2022, 2019; Nakadate et al., 2012; Roy et al., 2016; Ryan et al., 2015; Tanaka et al., 2018; Tyssowski et al., 2018; Watanabe et al., 1996; Yassin et al., 2010). We have now referenced some of those articles in the manuscript (Introduction). But more importantly, our claim is that the CSD is necessary to drive Fos. The fact that CSD can occur in a single hemisphere allows, within a brain, to control for direct ECS response and phase III oscillations. In unilateral CSD, contralateral hemispheres displayed Fos levels in the cortex that were similar to sham mice. Investigating the exact plasticity pathway that is triggered by the CSD is an interesting question we are pursuing in follow-up work, but does not influence our conclusions here.

(2) Postictal EEG suppression is one of the most promising correlates of positive clinical outcomes after electroconvulsive therapy. Cortical SD is also tightly coupled with suppression of neuronal activity in affected regions. Although the authors report that postictal suppression is stronger after stimulations with cortical SDs than without SDs, the cortices affected (ipsi) and unaffected (contra) by unilateral cortical calcium/SD events exhibit identical suppression (Figure 6F). The result contradicts established knowledge in the field. If the calcium events are cortical SDs, they should induce EEG suppression only in the affected hemisphere.

The reviewer is correct in that postictal suppression is currently one of the best correlates of positive clinical outcome used in the clinic. However, note that EEG-based correlates are generally weak predictors of treatment outcomes: even recently identified correlates are unreliable between patients cohorts and account for less than 60% of the clinical outcome (Francis-Taylor et al., 2020; Scangos et al., 2019). See also our response to comment (2) of reviewer 3 on this topic.

The primary problem in the interpretation of postictal suppression of EEG activity is that it is unclear what the source of the EEG signal is in this case. During phase III oscillation following the CSD, cortex is silent and all of the EEG is likely driven by thalamic input to cortex. Note, source localization in EEG does not help as the source of the signal is likely the thalamic axons in cortex (Huels et al., 2023) (or their postsynaptic, and in this case subthreshold, effects). Whatever the source of the cortical surface EEG may be, we know that following CSD it cannot be cortical activity, hence, whatever is driving postictal suppression is also not cortical. If we had to speculate, postictal suppression is likely an exhaustion of thalamus in attempting to drive cortex without getting any excitatory feedback. During an epileptic seizure, thalamus oscillates at delta frequencies and cortex responds at each peak of the oscillation, generating ictal spikes on the EEG (Meeren et al., 2002; Polack et al., 2009). The fact that ictal spikes are missing in an “optimal” phase III oscillation is consistent with our findings that the CSD completely silences cortex for minutes. Thus, during an ECS-driven phase III oscillation, cortex does not respond to thalamic input and at some point thalamus runs out of energy to maintain the oscillation - it is likely the combination of thalamus running out of energy and a silent cortex that gives rise to postictal suppression. Note we do find an asymmetry for the mid-oscillation amplitude (Figure 6A), which in this interpretation can be explained by the thalamo-cortical circuit attempting to compensate for the absence of cortical feedback by increasing its drive to the CSD-affected hemisphere.

(3) The study states a beneficial role of calcium/SD waves in ECS effects. However, SD alters numerous aspects of brain function, leading to a range of effects that can underlie side effects as well. Assessment of the behavioral effects of stimulation with and without calcium/SD waves can help clarify the issue.

This is a misunderstanding. We claim (and show) that the CSD is necessary to drive immediate early gene expression following ECS. Based on this we speculate that the CSD “may serve as a more relevant biomarker for predicting and optimizing therapeutic outcomes of ECT” than EEG biomarkers.

We do not “state [that there is] a beneficial role of CSD” – we actually address the possibility that SD may not be therapeutically beneficial (Discussion). The suggestion of investigating behavioral consequences of CSD in mice is interesting, but likely not the most relevant follow-up work. This is for two reasons:

(1) Very different from humans, most mouse behavior does not depend on cortex (Kawai et al., 2015; Pandey et al., 2026). Thus, any potential behavioral consequences of ECS in mice are difficult to interpret in the context of clinical relevance. We don’t think there is any known biomarker based on mouse behavior that has any direct translational value to psychiatric treatments. The reviewer may disagree with this assessment but just consider that no mouse behavior assay has ever been instrumental to the development of a novel psychiatric treatment.

(2) The more direct – and simpler approach – is to directly test whether CSD correlates with treatment outcomes in patients. Our primary aim with this paper is to inspire exactly this. Note, we are currently pursuing this as well, of course. Monitoring CSDs in humans during ECT is likely possible with fNIRS or other hemodynamic-based measurements (Discussion).

However, to reiterate – the claim of the manuscript is exactly as stated in the title. Thus, demonstrating the clinical relevance of the CSD is outside of the scope of the current manuscript, but will be trivial to prove by measuring treatment outcomes while routinely monitoring patients for CSD.

The results of the work suggest that cortical SD can contribute to electroconvulsive therapy-related mechanisms and help to optimize the stimulation parameters to achieve maximal therapeutic effect.

Reviewer #1 (Recommendations for the authors):

The results of sham tests are shown only for the immunohistochemical data. However, results from control experiments should be provided for the EEG and calcium data.

Sham data are included in Author response image 1. We are not sure how they support our conclusions and have left them out of the manuscript. A sham stimulation just characterizes ongoing activity under anesthesia. The more direct comparison is the difference between pre- and post-ECS activity. Baseline widefield calcium imaging data is already shown in Figure 1, 2, 6; baseline mouse EEG data is already shown in Figure 1 and 6. Baseline EEG in patients is not available in the present dataset.

Sham ECS does not result in phase III oscillations or a CSD.

(A) Representative spectrograms (top) and raw traces (bottom) of concurrent EEG (left) and widefield calcium imaging (right) during a sham ECS session. (B) Population raster plot (top) and two example activity traces of neurons (bottom) during a sham ECS two-photon recording.

Moreover, baseline (pre-ECS) calcium and EEG activity should be shown in Figures 1 and 6 to correctly assess the changes induced by stimulation.

We are not sure we understand as pre-ECS data are already shown in Figures 1 and 6. We assume the reviewer may mean “more baseline data should be shown”? We have extended the time scale of the relevant panels in Figure 1, 2 and 6 to include 30 s instead of 10 s of pre-ECS data.

Using the term 'oscillations/phase III oscillations' instead of 'seizures' throughout the text is confusing because the word covers a wide range of brain oscillations - from normal to pathological ones.

This is indeed confusing – but the confusion arises from the often imprecise usage of the term “seizure” in the ECT literature. The EEG response to ECS is not equivalent to that of an epileptic seizure. The description of a “phase III oscillation” is a more precise description of the EEG signature. It was introduced by (Brumback and Staton, 1982) and is not our terminology. We dedicate an entire paragraph in the introduction to this distinction. We are not sure how to make this clearer in the current manuscript. Continuing to describe the EEG response to ECS as “seizure” is inaccurate, and mechanistically misleading, especially given that cortex is silent during the phase III oscillation following the CSD (i.e. cortex cannot be “seizing” during this phase of the EEG response, see our point above on postictal EEG suppression).

The presentation of clinical data is scarce and unclear. e.g., the authors claim that the oscillation frequency is similar in mice and humans (lines 206-208). However, in Figure 1F, oscillations during the early post-ECS phase have twice the frequency (6-7 Hz) in the patient EEG recording (6-7 oscillations per 1 second) compared to the mouse EEG (3 Hz, i.e., 6 oscillations per 2 seconds during 73-75 s). It seems a bit odd because Figure 1H shows that the frequency does not exceed 5 Hz, even in patients.

Please excuse, this is our mistake. The x-axis label in the human data of Figure 1F was incorrect. It should have read 1 to 3 s, not 1 to 2 s. Window sizes were of course matched between mice and human data and are all 2 s. The mistake is now corrected. The frequencies in patients and mice are compared in Figure 1H and are not different.

A part of the discussion (lines 627-635) is based on factual inaccuracy: cortical SD cannot invade the hippocampus in the in vivo brain (only in slices), although SD can occur in the hippocampus in response to generalized seizures.

It would be helpful if the reviewer would back up this claim with references. Short of this we are left to speculate - we suspect the reviewer may be referring to earlier work in the rat cortex, that claimed that a cortical SD can only invade hippocampus if glia has been impaired (Largo et al., 1997). This is now an outdated model: there are more recent reports of CSD invading the hippocampus (Bahari et al., 2020; Bonaccini Calia et al., 2022). If the reviewer has specific concerns with any of these papers, we would be happy to discuss in more detail, but the reviewers’ claim seems unfounded here.

It is reasonable to expect that bilateral stimulation produces bilateral calcium/SD waves. In Rosenthal's experiments, this situation was most common. In the present study, bilateral ECS triggers mostly unilateral calcium waves. Do you have any idea what the reasons for the result are? As stimulation parameters (polarity, intensity, and frequency) have been shown to control the occurrence of SD waves, their unilateral pattern suggests non-uniform stimulation conditions. I am curious whether the uni- or bilateral pattern of calcium/SD waves depends on the ECS parameters.

The main difference between our patient and mouse data is the polarity of stimulation. For patient data, the polarity of the current alternates with every pulse, while in our mouse data the stimulation was always right unipolar. This is discussed when we mention the asymmetry of SD in our data (Results, Methods) - we suspect the reviewer may have missed this.

Reviewer #2 (Public review):

Summary:

This manuscript addresses the question of mechanisms underlying the therapeutic effects of electroconvulsive therapy (ECT). Clinical efficacy of ECT in major depression (and other disorders) is well established and has often been assumed to be a direct consequence of seizure activity generated by the current application. However, as the authors point out, this explanation is unsatisfactory. A recent study (Rosenthal et al., 2025) provided evidence that ECT generates a wave of cortical spreading depolarization (CSD) in mice, and initial evidence that similar events were generated in patients undergoing ECT. Based on their observations, Rosenthal et al. proposed that CSD, rather than seizure, may engage plasticity mechanisms that contribute to the brain's clinical response to ECT. The current study adds to that prior work by reporting other consequences of CSD, in addition to sustained Ca2+ elevations. The current study also links EEG characteristics immediately following the ECT with the likelihood of generating a CSD, which can help optimize ECT parameters.

Strengths:

An important research topic, linking a large set of rodent studies with a limited clinical EEG data set.

The data acquisition and analyses appear to be of very high quality, and the main results are well illustrated.

Association between EEG characteristics linked to good clinical outcome matched by mouse EEG data linked to CSD.

Characterization of multiple consequences of CSD following ECT in the mouse brain.

Weaknesses:

(1) The main characterization of CSD propagation comes from GCaMP Ca2+ measurements, as previously reported (Rosenthal et al., 2025). That prior study also provided key electrophysiological evidence of CSD with a DC shift after ECT in mice (supplemental data). Given the prior evidence for ECT-CSD, the additional measures shown in the current manuscript are fully expected. Thus, the 2-photon imaging of Ca2+ elevations following CSD (Figure 4) is consistent with prior 2-photon imaging studies of CSD, and the complex hemodynamic and pH changes are expected to contribute to propagation of EGFP fluorescence changes (Supplemental Figure 5). These data are well presented, but, contrary to the results section here, these results appear confirmatory rather than necessary to build a case that the key event generated by ECT is a CSD.

This is correct, our work confirms that the calcium event following ECS is a CSD. The main claim of our paper is that ECS drives immediate early gene expression in a CSD-dependent manner.

However, note that the Rosenthal paper concluded that the calcium event is a CSD while providing only little evidence for that claim – mind you, we agree with their interpretation, but provide more evidence for the conclusion. We are happy to discuss the limitations of the Rosenthal paper as highlighted below more prominently in the manuscript, if the reviewer thinks this would be helpful, but we think that is likely not necessary.

Briefly, in the data presented in (Rosenthal et al., 2025), the only support for the calcium response being a CSD is the speed of propagation and the DC shift reported in extracellular recordings. We add to this by showing that the spread of the calcium event follows the pattern expected by a CSD through cortical layers (Figure 4), results in heterogeneous returns to baseline calcium levels (Figure 4), causes vasoconstriction (Figure S5), travels at the speed expected of a CSD regardless of stimulation parameters (Figure S3) and causes Fos expression (Figure 5).

Most importantly, the ECS as used by Rosenthal and colleagues is not a mouse model of ECT, in the sense that is not a scalp electrical stimulation. The stimulation method is fundamentally different between our two articles: (Rosenthal et al., 2025) implanted stimulating electrodes directly in the mice’s dorsal cranium. Direct cortical stimulation is well known to be able to cause CSD (Leao, 1944). However, it is unclear whether direct cortical stimulation is a useful model for ECT. We suspect that Rosenthal and colleagues were led to believe that auricular stimulation does not work because they saw no evidence of a cortical seizure in calcium recordings following stimulation. (See discussion on the confusion of phase III oscillation and seizure in the ECT literature.) We suspect that EEG recordings following direct cortical stimulation would reveal a very different EEG pattern from that observed in patients. This highlights the importance and novelty of the comparison of mouse and human EEG we present in Figure 1. Note, in our auricular stimulation preparation we do not observe any seizure-like activity in cortex that lasts beyond the stimulation (compare Rosenthal’s Figure 1E vs Figure 1F here). Given the EEG similarity we show between mice and patients, we suspect the cortical seizure Rosenthal and colleagues find is a methodological artifact. This is puzzling to us, as Rosenthal and colleagues do briefly mention a single mouse example with an extracellular electrophysiology recording compatible with CSD following auricular stimulation (Supplementary Figure 2).

Thus, not only is it necessary to add evidence to the interpretation that the ECS-driven calcium event is indeed a CSD, but also that it can be triggered by a stimulation method that successfully replicates the known EEG response of human patients.

(2) The authors state that "our conclusion that CSD is the primary driver of plasticity is based on its role in driving Fos expression" (line 472). Related to the point above, there is already a very well-established literature showing that CSD leads to rapid and robust Fos expression in rodent cortex, so this is fully consistent with prior work. The prior work, CSD-fos work, should be summarized and/or cited more clearly in the manuscript. Showing that Fos increases only in the hemisphere where there is a large CSDCa2+ wave is a clear demonstration of this. While Fos increases can certainly be well linked to plasticity in some experimental paradigms, the implication that Fos increases underlie CSD-induced plasticity and possibly therapeutic effects of ECT is not appropriate. Fos increases after CSD are a reliable marker of the very strong neuronal activation that occurs, but Fos increases are not specific for plasticity and can be activated by challenges that do not generate synaptic plasticity. A range of other gene expression changes have been identified with CSD and may contribute to adaptive plasticity; these could be mentioned alongside speculation about Fos. To support the main conclusions of this paper about CSD driving plasticity via Fos, Fos knockout or knockdown studies are needed, as has been used in prior plasticity studies.

We have added additional references to the CSD-Fos literature in the discussion. Regarding the role of Fos as a marker of plasticity rather than activity, we discuss this point in our reply to comment (1) of reviewer 1. Concerning the expression of other genes, we already referenced the TRKB/BDNF pathways (Discussion). We have now added references on RNA-seq following CSD, which show that Fos is one of the most differentially expressed genes following CSD (Dell’Orco et al., 2023). Regarding the use of Fos knockout/knockdown lines, please see our reply to the reviewer’s comment (4).

Reviewer #2 (Recommendations for the authors):

(3) The Results and Discussion sections should be revised to better reflect the prior discovery of CSD following ECT in rodents and initial evidence in humans, as discussed in the first point in the Weaknesses section above.

The prior discovery of the fact that ECS can drive CSD is first mentioned in the third sentence of the abstract “However, this view is challenged by the recent finding that electroconvulsive stimulation (ECS) can trigger a cortical spreading depression (CSD).” (The abstract has no references, but the introduction should make it clear what is meant).

There is an entire paragraph of the introduction discussing the Rosenthal results.

The first time we discuss our results (end of the Introduction) we say: “Consistent with previous work (Rosenthal et al., 2025), we observed a slow travelling calcium event that appeared to be a CSD.” The Rosenthal paper is cited in 8 times in total throughout the manuscript.

We are unsure what the reviewer is asking us to do here. As mentioned above, if any revision is warranted regarding that reference, it should be to clarify that the Rosenthal paper used direct intracranial stimulation rather than ECS, and did not fully confirm the calcium event as a CSD – but they should be credited for finding the first preliminary evidence for CSD following ECS in patients.

(4) To support the authors' statement that "CSD is the primary driver of plasticity is based on its role in driving Fos expression", additional experiments with Fos knockdown or knockout (or alternative interventions) are needed.

Our main claim is that CSD causes Fos expression in the cortex following ECS. The argument that Fos can be used as a marker of plasticity follows from the literature, not from the experiments done here – this would require a form of functional plasticity measurement, which is outside the scope of this paper (and might not be the most relevant direction, see our reply to comment (3) of reviewer 1). The only observation that a Fos genetic manipulation would give us is the lack or reduction of Fos expression following CSD, which would be orthogonal to the points we make here.

(5) It would be helpful to use a more specific term than "Ca2+ event" in Figure 2D and throughout the related Results section. It is assumed that this is the large propagating Ca2+ event attributed to CSD, but the terminology is important, as all the other events in the recording (including during Pre-ECT and the Direct ECT period) are also Ca2+ events.

We have added a clear definition of what we mean on first usage (Results). The reason to call it a calcium event, and not a CSD, is that we did not want to jump to conclusions. We do think that the event is a CSD, but conclusive proof of that is still lacking (in both our work and that of Rosenthal). It is conceivable that the event propagates via a different mechanism than a CSD.

(6) The authors should comment on differences among rodent models of ECT stimulation, especially with direct and ear clip methods, as discussed in the context of translational value (Theilmann et al., 2014).

The Theilmann 2014 paper compares ECS delivered via auricular stimulation and intracranial electrodes. They compare the two stimulation methods, but use stimulation currents, total charges, and stimulation duration that were not matched. In the case of stimulation current, those used in auricular stimulation are approximately 8 fold higher than what they use for intracranial stimulation. Moreover, the stimulations were performed in awake rats. This is scientifically - and ethically, even for 2014 - questionable in light of the fact that this is aimed at developing a model for ECT, which is always done under general anesthesia. They conclude that cortical stimulation has less adverse effects and is more effective in reducing immobility in a forced swim test. The confound in the interpretation of these results is that all stimulation was performed in awake animals. The reason this is no longer done in humans is that it is extremely painful. Direct cortical stimulation is likely much less painful (for the same reason TMS is less painful). This would explain their findings of increased adverse effects with auricular electrodes. Given the differences in stimulation parameters used, the difficulty of calibrating equivalent doses of auricular and intracortical stimulations, as well as the small effect sizes reported, we don’t think the results allow for any solid conclusions as to which method is more effective in reducing immobility in a forced swim test. However, even if one would assume the intracortical ECS is more ‘effective’, this is hardly relevant, as we are interested in using mouse ECS as a model for human ECT. One could speculate that intracranial ECS might also exhibit higher clinical benefit than surface ECS in patients, but that is not the scope of our research, and probably not clinically relevant. Finally, while the authors do include EEG recordings in the rat – these were not compared to patient recordings, and from visual inspection do not resemble patient EEG recordings that we have seen. We would argue that the best rodent model of ECT stimulation is the one that triggers neuronal activity most similar to that observed in patients. We have added a brief discussion of these points to the corresponding Methods section.

Reviewer #3 (Public review):

Summary:

This manuscript combines widefield calcium imaging, electroencephalography, 2-photon imaging, and immunohistochemistry in mice to re-demonstrate that electroconvulsive stimulation (ECS) induces a seizure followed by cortical spreading depolarization, as previously shown. The putative novel finding - which is not unexpected - is that ECS is also correlated with increased expression of the immediate early gene cFOS, although this has also been shown previously. The authors speculate that CSD drives cFOS expression, which might contribute to the therapeutic effects of ECT; however, experiments performed do not provide causal evidence for this hypothesis. Instead, the authors use expression of cFOS - a nonspecific activity-dependent gene induced in various pathological and non-therapeutic contexts - as a proxy for plasticity and/or therapeutic effect. Hence, overall, the significance of the findings is limited and primarily serves to replicate prior work, with the evidence evaluated as incomplete.

Strengths:

The experiments are generally well executed from a technical perspective.

Main Weaknesses to be addressed in revision:

(1) The main findings of this paper are replication experiments of prior work, and thus, the novelty and significance of this manuscript are relatively limited.

This appears incorrect. It was known that direct cortical stimulation (as was done in the Rosenthal paper) can drive a calcium event that resembles a CSD. It was also known that CSD can drive Fos expression. What was not known is that the Fos expression driven by ECS is fully explained by the calcium event (putative CSD). This is particularly relevant as most people still erroneously assume Fos is a marker or neuronal activity, and prior work has come to the conclusion that ECT does not drive, but likely downregulates Fos expression (Calais et al., 2013; Morinobu et al., 1995; Park et al., 2014; Winston et al., 1990). We have added a more prominent discussion section on this point.

- It is already known that the mean frequency of ECT-induced seizures decays between peak and offset in humans (Stuiver et al. Clin Neurophysiol. 2026 Jan:181:2111439. doi: 10.1016/j.clinph.2025.2111439) and mice (Murakami et al. J Pharmacol Sci 2008 Jan;106(1):78-83. 10.1254/jphs.FP0071453), which the authors re-demonstrate in Figure 1.

The importance of Figure 1 is to demonstrate that ECS delivered with auricular electrodes in mice causes an EEG signature that is very similar to that seen in patients. We do not claim we are the first to describe characteristics of phase III oscillations in either patients or mice (we have added the Murakami reference to the manuscript). The reply to comments (1) and (6) of reviewer 2 highlights why this comparison is so important – it was not done in the Rosenthal paper the reviewer mentions below, and it is not clear whether the intracortical stimulation used there even drives a comparable EEG response (given the calcium activity shown, we suspect the answer is no). To the best of our knowledge this direct comparison is novel – but again the key novelty of our work we highlight is the one described in the title.

- It has already been demonstrated that ECT in mouse models induces lateralized CSD waves in a manner that depends on stimulation parameters and the initial evoked response during stimulation (Rosenthal et al. Nat Comm. 2025 May 18;16(1):4619. doi: 10.1038/s41467-025-59900-1); the authors replicate this in Figures 1, 2, 3, 6.

It has indeed been demonstrated that ECS delivered using intracranial electrical stimulation can trigger CSD-like events (e.g. Rosenthal et al.). However, the fact that localized intracranial electrical stimulation can trigger a CSD has been shown quite a while ago already (see e.g. (Leao, 1944)). This is not the case for surface stimulation the way it is done in ECT and the way we do it. Nevertheless, note we give full credit to the Rosenthal work for making this connection. Our main contribution – as highlighted by the title – is showing that the Fos expression driven by ECS is fully explained by the CSD.

- It is already widely established that EEG and calcium signals are highly concordant in mouse brain physiology, as shown in Figure 1.

If the reviewer has references for this claim, we would be happy to add to the manuscript – we are not aware of any such work. As far as we are aware, this is still an area under active investigation – calcium signals correlate (locally) strongly with shank recordings (Wei et al., 2020), but how this translates into an EEG signal is speculative.

It is already known that CSD propagates from supragranular to granular and infragranular layers (Zakharov et al. Epilepsia. 2019 Dec;60(12):2386-2397. doi: 10.1111/epi.16390) as shown in Figure 4.

The reviewer may be jumping to conclusions here. It is correct that this has been shown for a CSD. The more important question (and the reason we did this experiment) is whether the calcium event triggered by ECS is indeed a CSD. We try to be careful to describe it as a calcium event in the results (mind you the calcium event in the Rosenthal is very likely a CSD as their intracortical stimulation (‘ECS’) is likely equivalent to the electrical stimulations used in the discovery of the CSD). We then perform a series of comparisons to see whether the calcium event has the known characteristics of a CSD – and we conclude everything we test is consistent with it being a CSD. Note, once again, that we do not claim novelty in any of this – the primary novelty is the link between ECS, Fos and CSD.

- It is already known that CSD waves induce cFOS expression (e.g., Dell'Orco et al. Front Cell Neurosci. 2023 Dec 14:17:1292661. doi: 10.3389/fncel.2023.1292661; Hermann and Hossman. Neuroscience. 1999 Jan;88(2):599-608. doi: 10.1016/s0306-4522(98)00249-8) as the authors replicate in Figure 5.

That is correct. The question however is how much of the Fos expression is explained by CSD. Prior work that has looked at Fos expression in response to ECS has found that ECS results in a slight reduction of Fos expression (Calais et al., 2013; Park et al., 2014). We suspect this is the result of not triggering a CSD. We do not claim to have discovered that CSD induces Fos expression. The novel contribution, which is the main claim of the paper, is that in the context of ECS the entirety of the cortical Fos expression can be explained by the CSD. This links the relative contributions of multiple components of the ECS response to a known marker of neuronal plasticity.

Minimally, the authors should revise claims regarding novelty, as the manuscript, as written, is misleading to a reader not familiar with the field. There is limited innovation in re-demonstrating that these events are seizures and that they involve spreading depolarization.

There is probably a misunderstanding here. We argue and show that there is no cortical seizure following ECS – that is why we refer to the EEG responses as phase III oscillations (characteristic of a silent cortex). And there is little prior evidence that the calcium events triggered by ECS are indeed a CSD (we think this is likely the case, but demonstrating this conclusively will require further work). If the reviewer has references for this, we would be happy to discuss. Note, the Rosenthal et al. paper just assumes (probably correctly) that they are a CSD, but does not demonstrate this. We don’t fully demonstrate this either, we just provide additional evidence. But once again, the novelty is in the title of the manuscript, and we do not claim any other novelty to the best of our reading of our manuscript. If the reviewer has a particularly misleading passage in mind, we are happy to rephrase.

(2) The authors frame their hypothesis that CSD could be a potential mediator of the therapeutic effects of ECT, but they do not measure therapeutic effects or directly test this hypothesis. The principal advancement of the paper is showing that ECT-induced CSD triggers hemisphere-specific cFOS expression as a proxy of plasticity. However, it is already known that CSD induces cFOS expression (as noted above). The observation that cFOS expression was induced only by CSD, not by the initial seizure, is likely a byproduct of the greater activity induced by CSD than by seizure. cFOS expression is nonspecific to plasticity or therapeutic effects and can be triggered by many non-therapeutic interventions. The cFOS data thus do not meaningfully measure therapeutic plasticity. The authors also selectively cite references suggesting that EEG metrics such as seizure duration predict positive therapeutic outcomes, but this link is controversial and not well established in the clinical literature.

We are not sure what the reviewer means by “cFOS expression is nonspecific to plasticity”. Does the reviewer mean Fos has other roles beside driving neuronal plasticity? That is very likely correct, but it is unclear how that is relevant. Fos is a key driver of a number of neuronal plasticity pathways (Chaudhuri et al., 2000; Cohen and Greenberg, 2008; Cruz et al., 2015; Durchdewald et al., 2009). Which exact pathways are driven by ECS is an interesting question that we are currently pursuing in follow-up work. Given what we know about Fos expression, it is probably well within reasonable bounds to conclude that Fos increases result in neuronal plasticity. Fos expression directly drives network plasticity (Yap et al., 2021): "our findings indicate that Fos expression has an instructive role in orchestrating persistent circuit modifications”. Likewise, Fos induction is strictly required for experience-dependent representational plasticity during learning (de Hoz et al., 2018): “locally blocking c-Fos expression caused […] decreased cortical experience-dependent plasticity, without affecting baseline excitability or basic auditory processing.”. Calcium activity explains about 15% of the variance of Fos expression (Mahringer et al., 2022). This is likely driven by the correlation between activity and plasticity, not by a direct necessity for Fos expression to maintain neuronal activity. This is consistent with the finding that Fos as a transcription factor does not function to maintain spiking activity; it is part of the gene-regulatory machinery that converts patterned synaptic input into lasting plastic change. Fos is induced by NMDA/Ca2+, ERK, and CREB/Elk signaling rather than by firing alone (Fields et al., 1997; Xia et al., 1996), and those same pathways are required for the transcriptional program that stabilizes long-term potentiation and other durable synaptic modifications (Davis et al., 2000). As an AP-1 transcription factor, Fos drives downstream gene expression, so its appearance is better read as entry into a plasticity-related nuclear program than as a measure of ongoing excitability (Minatohara et al., 2015; Morgan and Curran, 1991; Sheng and Greenberg, 1990). That interpretation is consistent with our work showing that early Fos expression preferentially marks neurons that later undergo the strongest learning-related functional changes (Mahringer et al., 2019) and tracks functional reorganization during learning rather than simple recent activation (Mahringer et al., 2019).

The link between CSD and therapeutic effect is a speculation we make in the manuscript, not a conclusion. We are of course in the process of performing follow-up work to test whether CSD in patients is a better predictor of treatment outcome than EEG based metrics – no experiment we can do in mice will be able to test the hypothesis that CSD is the mediator of the therapeutic benefit of ECT.

Regarding the power of EEG metrics to predict therapeutic outcomes, we fully agree with the reviewer. The literature on the reliability of EEG metrics computed from phase III oscillation data is controversial and noisy – this is something we establish in the introduction to motivate our research into other biological processes that could explain how ECT works. Note however, this is certainly not a fringe view – see e.g. comment (2) of reviewer 1: “Postictal EEG suppression is one of the most promising correlates of positive clinical outcomes after electroconvulsive therapy“ We think the reason for this is that a CSD is necessary for therapeutic benefit, but only has minor effects on the phase III oscillation – note this is a hypothesis based on our results that is trivial to test in patients (which are in currently investigating).

Minor Weaknesses:

(3) For the n=3 mice used for concurrent 2P imaging with microprism implant, these animals also had ChrimsonR co-expression, but there are no optogenetic studies described in this paper, which is confusing. Yet, this co-expression introduces a significant confound, as GCaMP6 emission (525/50nm band in this study) will overlap substantially with the ChrimsonR excitation spectrum. Thus, the fluorescence emission used to image these neurons may be optogenetically activating them at the same time. Please explain.

Whenever possible, we use mice for multiple experiments in the lab. This is done to reduce the total number of mice used for experiments for ethical reasons. For the mice in question, ChrimsonR was injected in the retrosplenial cortex (Methods), which was originally done to stimulate locally the axons projecting in the imaging area (primary visual cortex here). Thus the labelling is very sparse, and perfectly compatible with two-photon GCaMP6f imaging. Fluorescence emission is far too weak to activate ChrimsonR.

Qualitatively, one can mentally compare the light power we use to activate optogenetic tools, which tends to be blindingly bright (one shouldn’t look into the optogenetic stimulation laser), with the fluorescence emission from two-photon imaging of calcium indicators, which tends to be barely visible by eye.

Quantitatively, one can estimate this as follows: At 510nm emission, each photon carries an energy of about E=hcλ=3.9∗10−19J. Assuming a neuron that strongly expresses GCaMP6f under two-photon excitation would emit a very high 107photons per second (Har-Gil et al., 2018), its total emission power would be P = Nphotons * E ≈ 4pW. Assuming this is spread over the surface of the cell (sphere with 10 µm diameter), this translates to E=hcλ=3.9∗10−19J. This is several orders of magnitude below the value of 1 mW/mm2 irradiance required to activate ChrimsonR modestly at peak absorption, which is 80nm away from GCaMP6f emission (Klapoetke et al., 2014). Note that this would be true even when ChrimsonR is injected at the site of imaging (Vasilevskaya and Keller, 2026).

(4) Incision of the cortex for implantation of a prism is a significant cortical injury that likely induces CSD instantaneously and may change the propensity for CSD in subsequent recordings. Please comment on this limitation and address how much time elapsed after surgery before imaging.

We suspect that the question is driven by a misunderstanding. While it is likely that the implantation triggers a CSD (and likely so does a standard two-photon window implantation), the implantation surgery and the experiments/imaging are separated by at least 3 weeks (Methods). We have never observed spontaneous CSDs in the days and weeks following an implantation surgery.

(5) Method details are missing or insufficiently described for location, titer, and injection strategy for 2-photon experiments.

We have added additional details as requested by the reviewer (Methods).

(6) Given the wide range of parameters used for ECS in mice and ECT in humans, the authors should provide tables for what stimulation parameters were used for each recording. These protocols were chosen manually rather than randomly or systematically, which introduces confounding factors into analyses that use parameters as an independent variable.

We have added two tables (Table S3 and Table S4) that displays the stimulation parameters used for each figure, as well as the distribution of parameters for mice and patients. More importantly, the properties of the travelling calcium event do not depend on the stimulation charge (Figure S3), which removes this confounding factor and supports the idea that the calcium event is a CSD.

(7) While much of the cFOS staining after unilateral CSD shows hemisphere-specific asymmetry, several regions (piriform cortex, amygdala, thalamus) do appear to have bilateral cFOS expression. Please comment on this.

That is correct – only the cortical expression of Fos depends on the cortical CSD (see our reply to comment (1) of reviewer 1). Quantification of the whole-brain Fos expression following ECS is outside the scope of the manuscript, but it is something we are currently pursuing for separate publication. We have now reworked the section describing the Fos expression to make it clear that we are only talking about cortical expression of Fos (Results).

(8) The discussion states: "If CSD accounts for plasticity effects, triggering a CSD in a non-seizure context may be sufficient to elicit therapeutic effects. This is supported by the clinical success of ultra-brief stimulation treatments that do not cause seizures, such as rTMS with accelerated protocols, which achieves treatment efficacy on par with ECT for major depressive disorder". Are the authors implying that TMS induces CSD? What evidence supports this idea?

That was indeed our speculation based on ongoing work on TMS in the lab – but it was poorly phrased and unnecessary. We have rephrased.

(9) This statement - "Assuming psychosis is the result of thalamocortical coupling that is too weak in frontal areas of the cortex" (lines 583-585) - may be overly speculative.

It is speculative indeed – but the speculation is not unfounded and has been made previously. The primary evidence is correlative in that schizophrenia is characterized by a reduction in coupling between thalamus and frontal areas of cortex – see e.g. (Giraldo-Chica and Woodward, 2017; Vinogradov et al., 2023). We have argued in previous work that combining this with computational models of psychosis (Sterzer et al., 2018), it is not unreasonable to speculate that a reduction in the coupling between thalamus and frontal areas of cortex could explain psychosis (Keller and Sterzer, 2024). The value of that speculation here is that it forms a testable hypothesis for the mechanism of action of ECT. Nevertheless, we have rephrased the statement slightly to make it clearer that this is still speculation.

Reviewer #3 (Recommendations for the authors):

The authors should design/execute an experiment(s) to attempt to prove causality between CSD, calcium influx, cFOS expression, and therapeutic effect. At a minimum, the authors need to revise claims regarding novelty, as the manuscript, as written, is misleading to a reader not familiar with the field.

Regarding novelty – see discussion above.

Regarding therapeutic effects – we are in the process of testing this in patients. Using CSD measurements during ECT to test whether CSD is a better predictor of treatment outcome. I suspect that this will, however, require many more labs to come to firm conclusions. Our aim here is to inspire these experiments. That is why we speculate about clinical relevance in the abstract and the discussion.

References

Bahari, F., Ssentongo, P., Liu, J., Kimbugwe, J., Curay, C., Schiff, S.J., Gluckman, B.J., 2020. Seizure-associated spreading depression is a major feature of ictal events in two animal models of chronic epilepsy. https://doi.org/10.1101/455519

Bolhuis, J.J., Hetebrij, E., Den Boer-Visser, A.M., De Groot, J.H., Zijlstra, G.G.O., 2001. Localized immediate early gene expression related to the strength of song learning in socially reared zebra finches. European Journal of Neuroscience 13, 2165–2170. https://doi.org/10.1046/j.0953816x.2001.01588.x

Bonaccini Calia, A., Masvidal-Codina, E., Smith, T.M., Schäfer, N., Rathore, D., Rodríguez-Lucas, E., Illa, X., De la Cruz, J.M., Del Corro, E., Prats-Alfonso, E., Viana, D., Bousquet, J., Hébert, C., Martínez-Aguilar, J., Sperling, J.R., Drummond, M., Halder, A., Dodd, A., Barr, K., Savage, S., Fornell, J., Sort, J., Guger, C., Villa, R., Kostarelos, K., Wykes, R.C., Guimerà-Brunet, A., Garrido, J.A., 2022. Full-bandwidth electrophysiology of seizures and epileptiform activity enabled by flexible graphene microtransistor depth neural probes. Nat. Nanotechnol. 17, 301–309. https://doi.org/10.1038/s41565-021-01041-9

Brumback, R.A., Staton, R.D., 1982. The Electroencephalographic Pattern during Electroconvulsive Therapy. Clinical Electroencephalography 13, 148–153. https://doi.org/10.1177/155005948201300306

Calais, J.B., Valvassori, S.S., Resende, W.R., Feier, G., Athié, M.C.P., Ribeiro, S., Gattaz, W.F., Quevedo, J., Ojopi, E.B., 2013. Long-term decrease in immediate early gene expression after electroconvulsive seizures. J Neural Transm (Vienna) 120, 259–266. https://doi.org/10.1007/s00702-012-0861-4

Chaudhuri, A., Zangenehpour, S., Rahbar-Dehgan, F., Ye, F., 2000. Molecular maps of neural activity and quiescence. Acta Neurobiol Exp (Wars) 60, 403–410. https://doi.org/10.55782/ane-2000-1359

Cohen, S., Greenberg, M.E., 2008. Communication between the synapse and the nucleus in neuronal development, plasticity, and disease. Annu Rev Cell Dev Biol 24, 183–209. https://doi.org/10.1146/annurev.cellbio.24.110707.175235

Cruz, F.C., Javier Rubio, F., Hope, B.T., 2015. Using c-fos to study neuronal ensembles in corticostriatal circuitry of addiction. Brain Res 1628, 157–173. https://doi.org/10.1016/j.brainres.2014.11.005

Davis, S., Vanhoutte, P., Pages, C., Caboche, J., Laroche, S., 2000. The MAPK/ERK cascade targets both Elk-1 and cAMP response element-binding protein to control long-term potentiation-dependent gene expression in the dentate gyrus in vivo. J Neurosci 20, 4563–4572. https://doi.org/10.1523/JNEUROSCI.20-12-04563.2000

de Hoz, L., Gierej, D., Lioudyno, V., Jaworski, J., Blazejczyk, M., Cruces-Solís, H., Beroun, A., Lebitko, T., Nikolaev, T., Knapska, E., Nelken, I., Kaczmarek, L., 2018. Blocking c-Fos Expression Reveals the Role of Auditory Cortex Plasticity in Sound Frequency Discrimination Learning. Cereb Cortex 28, 1645–1655. https://doi.org/10.1093/cercor/bhx060

Dell’Orco, M., Weisend, J.E., Perrone-Bizzozero, N.I., Carlson, A.P., Morton, R.A., Linsenbardt, D.N., Shuttleworth, C.W., 2023. Repetitive spreading depolarization induces gene expression changes related to synaptic plasticity and neuroprotective pathways. Front. Cell. Neurosci. 17. https://doi.org/10.3389/fncel.2023.1292661

Durchdewald, M., Angel, P., Hess, J., 2009. The transcription factor Fos: a Janus-type regulator in health and disease. Histol Histopathol 24, 1451–1461. https://doi.org/10.14670/HH-24.1451

Fields, R.D., Eshete, F., Stevens, B., Itoh, K., 1997. Action potential-dependent regulation of gene expression: temporal specificity in ca2+, cAMP-responsive element-binding proteins, and mitogen-activated protein kinase signaling. J Neurosci 17, 7252–7266. https://doi.org/10.1523/JNEUROSCI.17-19-07252.1997

Fleischmann, A., Hvalby, O., Jensen, V., Strekalova, T., Zacher, C., Layer, L.E., Kvello, A., Reschke, M., Spanagel, R., Sprengel, R., Wagner, E.F., Gass, P., 2003. Impaired Long-Term Memory and NR2AType NMDA Receptor-Dependent Synaptic Plasticity in Mice Lacking c-Fos in the CNS. J. Neurosci. 23, 9116–9122. https://doi.org/10.1523/JNEUROSCI.23-27-09116.2003

Francis-Taylor, R., Ophel, G., Martin, D., Loo, C., 2020. The ictal EEG in ECT: A systematic review of the relationships between ictal features, ECT technique, seizure threshold and outcomes. Brain Stimulation 13, 1644–1654. https://doi.org/10.1016/j.brs.2020.09.009

Giraldo-Chica, M., Woodward, N.D., 2017. Review of thalamocortical resting-state fMRI studies in schizophrenia. Schizophr Res 180, 58–63. https://doi.org/10.1016/j.schres.2016.08.005

Har-Gil, H., Golgher, L., Israel, S., Kain, D., Cheshnovsky, O., Parnas, M., Blinder, P., 2018. PySight: plug and play photon counting for fast continuous volumetric intravital microscopy. Optica 5, 1104. https://doi.org/10.1364/OPTICA.5.001104

Huels, E.R., Kafashan, M., Hickman, L.B., Ching, S., Lin, N., Lenze, E.J., Farber, N.B., Avidan, M.S., Hogan, R.E., Palanca, B.J.A., 2023. Central-positive complexes in ECT-induced seizures: Possible evidence for thalamocortical mechanisms. Clin Neurophysiol 146, 77–86. https://doi.org/10.1016/j.clinph.2022.11.015

Kawai, R., Markman, T., Poddar, R., Ko, R., Fantana, A.L., Dhawale, A.K., Kampff, A.R., Ölveczky, B.P., 2015. Motor cortex is required for learning but not for executing a motor skill. Neuron 86, 800– 812. https://doi.org/10.1016/j.neuron.2015.03.024

Keller, G.B., Sterzer, P., 2024. Predictive Processing: A Circuit Approach to Psychosis. Annu Rev Neurosci 47, 85–101. https://doi.org/10.1146/annurev-neuro-100223-121214

Kimpo, R.R., Doupe, A.J., 1997. FOS Is Induced by Singing in Distinct Neuronal Populations in a Motor Network. Neuron 18, 315–325. https://doi.org/10.1016/S0896-6273(00)80271-8

Klapoetke, N.C., Murata, Y., Kim, S.S., Pulver, S.R., Birdsey-Benson, A., Cho, Y.K., Morimoto, T.K., Chuong, A.S., Carpenter, E.J., Tian, Z., Wang, J., Xie, Y., Yan, Z., Zhang, Y., Chow, B.Y., Surek, B., Melkonian, M., Jayaraman, V., Constantine-Paton, M., Wong, G.K.-S., Boyden, E.S., 2014. Independent optical excitation of distinct neural populations. Nat Methods 11, 338–346. https://doi.org/10.1038/nmeth.2836

Largo, C., Ibarz, J.M., Herreras, O., 1997. Effects of the Gliotoxin Fluorocitrate on Spreading Depression and Glial Membrane Potential in Rat Brain In Situ. Journal of Neurophysiology 78, 295–307. https://doi.org/10.1152/jn.1997.78.1.295

Leao, A.A.P., 1944. Spreading depression of activity in the cerebral cortex. Journal of Neurophysiology 7, 359–390. https://doi.org/10.1152/jn.1944.7.6.359

Mahringer, D., Petersen, A.V., Fiser, A., Okuno, H., Bito, H., Perrier, J.-F., Keller, G.B., 2019. Expression of c-Fos and Arc in hippocampal region CA1 marks neurons that exhibit learning-related activity changes. https://doi.org/10.1101/644526

Mahringer, D., Zmarz, P., Okuno, H., Bito, H., Keller, G.B., 2022. Functional correlates of immediate early gene expression in mouse visual cortex. Peer Community Journal 2. https://doi.org/10.24072/pcjournal.156

Meeren, H.K.M., Pijn, J.P.M., Luijtelaar, E.L.J.M.V., Coenen, A.M.L., Silva, F.H.L. da, 2002. Cortical Focus Drives Widespread Corticothalamic Networks during Spontaneous Absence Seizures in Rats. J. Neurosci. 22, 1480–1495. https://doi.org/10.1523/JNEUROSCI.22-04-01480.2002

Minatohara, K., Akiyoshi, M., Okuno, H., 2015. Role of Immediate-Early Genes in Synaptic Plasticity and Neuronal Ensembles Underlying the Memory Trace. Front Mol Neurosci 8, 78. https://doi.org/10.3389/fnmol.2015.00078

Morgan, J.I., Curran, T., 1991. Stimulus-transcription coupling in the nervous system: involvement of the inducible proto-oncogenes fos and jun. Annu Rev Neurosci 14, 421–451. https://doi.org/10.1146/annurev.ne.14.030191.002225

Morinobu, S., Nibuya, M., Duman, R.S., 1995. Chronic antidepressant treatment down-regulates the induction of c-fos mRNA in response to acute stress in rat frontal cortex. Neuropsychopharmacology 12, 221–228. https://doi.org/10.1016/0893-133X(94)00067-A

Nakadate, K., Imamura, K., Watanabe, Y., 2012. Effects of monocular deprivation on the spatial pattern of visually induced expression of c-Fos protein. Neuroscience 202, 17–28. https://doi.org/10.1016/j.neuroscience.2011.12.004

Pandey, A., Kang, S., Pacchiarini, N., Wyszynska, H., Masseri, Z., O’Neill, J., Honey, R.C., Fox, K., 2026. Secondary Somatosensory Cortex Is Required for Learning but Not Execution of a Tactile Discrimination. Eur J Neurosci 63, e70390. https://doi.org/10.1111/ejn.70390

Park, H.G., Yu, H.S., Park, S., Ahn, Y.M., Kim, Y.S., Kim, S.H., 2014. Repeated treatment with electroconvulsive seizures induces HDAC2 expression and down-regulation of NMDA receptorrelated genes through histone deacetylation in the rat frontal cortex. Int J Neuropsychopharmacol 17, 1487–1500. https://doi.org/10.1017/S1461145714000248

Polack, P.-O., Mahon, S., Chavez, M., Charpier, S., 2009. Inactivation of the Somatosensory Cortex Prevents Paroxysmal Oscillations in Cortical and Related Thalamic Neurons in a Genetic Model of Absence Epilepsy. Cereb Cortex 19, 2078–2091. https://doi.org/10.1093/cercor/bhn237

Rosenthal, Z.P., Majeski, J.B., Somarowthu, A., Quinn, D.K., Lindquist, B.E., Putt, M.E., Karaj, A., Favilla, C.G., Baker, W.B., Hosseini, G., Rodriguez, J.P., Cristancho, M.A., Sheline, Y.I., William Shuttleworth, C., Abbott, C.C., Yodh, A.G., Goldberg, E.M., 2025. Electroconvulsive therapy generates a postictal wave of spreading depolarization in mice and humans. Nat Commun 16, 4619. https://doi.org/10.1038/s41467-025-59900-1

Roy, D.S., Arons, A., Mitchell, T.I., Pignatelli, M., Ryan, T.J., Tonegawa, S., 2016. Memory retrieval by activating engram cells in mouse models of early Alzheimer’s disease. Nature 531, 508–512. https://doi.org/10.1038/nature17172

Ryan, T.J., Roy, D.S., Pignatelli, M., Arons, A., Tonegawa, S., 2015. Memory. Engram cells retain memory under retrograde amnesia. Science 348, 1007–1013. https://doi.org/10.1126/science.aaa5542

Scangos, K.W., Weiner, R.D., Coffey, C.E., Krystal, A.D., 2019. An electrophysiological biomarker that may predict treatment response to ECT. J ECT 35, 95–102. https://doi.org/10.1097/YCT.0000000000000557

Sheng, M., Greenberg, M.E., 1990. The regulation and function of c-fos and other immediate early genes in the nervous system. Neuron 4, 477–485. https://doi.org/10.1016/0896-6273(90)90106-p

Sterzer, P., Adams, R.A., Fletcher, P., Frith, C., Lawrie, S.M., Muckli, L., Petrovic, P., Uhlhaas, P., Voss, M., Corlett, P.R., 2018. The Predictive Coding Account of Psychosis. Biol Psychiatry 84, 634–643. https://doi.org/10.1016/j.biopsych.2018.05.015

Tanaka, K.Z., He, H., Tomar, A., Niisato, K., Huang, A.J.Y., McHugh, T.J., 2018. The hippocampal engram maps experience but not place. Science 361, 392–397. https://doi.org/10.1126/science.aat5397

Tyssowski, K.M., DeStefino, N.R., Cho, J.-H., Dunn, C.J., Poston, R.G., Carty, C.E., Jones, R.D., Chang, S.M., Romeo, P., Wurzelmann, M.K., Ward, J.M., Andermann, M.L., Saha, R.N., Dudek, S.M., Gray, J.M., 2018. Different Neuronal Activity Patterns Induce Different Gene Expression Programs. Neuron 98, 530-546.e11. https://doi.org/10.1016/j.neuron.2018.04.001

Vasilevskaya, A., Keller, G.B., 2026. A functional influence based circuit motif that constrains the set of plausible algorithms of cortical function. eLife 15. https://doi.org/10.7554/eLife.110827.1

Vinogradov, S., Chafee, M.V., Lee, E., Morishita, H., 2023. Psychosis spectrum illnesses as disorders of prefrontal critical period plasticity. Neuropsychopharmacology 48, 168–185. https://doi.org/10.1038/s41386-022-01451-w

Watanabe, Y., Johnson, R.S., Butler, L.S., Binder, D.K., Spiegelman, B.M., Papaioannou, V.E., McNamara, J.O., 1996. Null Mutation of c-fos Impairs Structural and Functional Plasticities in the Kindling Model of Epilepsy. J Neurosci 16, 3827–3836. https://doi.org/10.1523/JNEUROSCI.16-1203827.1996

Wei, Z., Lin, B.-J., Chen, T.-W., Daie, K., Svoboda, K., Druckmann, S., 2020. A comparison of neuronal population dynamics measured with calcium imaging and electrophysiology. PLoS Comput Biol 16, e1008198. https://doi.org/10.1371/journal.pcbi.1008198

Winston, S.M., Hayward, M.D., Nestler, E.J., Duman, R.S., 1990. Chronic electroconvulsive seizures down-regulate expression of the immediate-early genes c-fos and c-jun in rat cerebral cortex. J Neurochem 54, 1920–1925. https://doi.org/10.1111/j.1471-4159.1990.tb04892.x

Xia, Z., Dudek, H., Miranti, C.K., Greenberg, M.E., 1996. Calcium influx via the NMDA receptor induces immediate early gene transcription by a MAP kinase/ERK-dependent mechanism. J Neurosci 16, 5425–5436. https://doi.org/10.1523/JNEUROSCI.16-17-05425.1996

Yap, E.-L., Pettit, N.L., Davis, C.P., Nagy, M.A., Harmin, D.A., Golden, E., Dagliyan, O., Lin, C., Rudolph, S., Sharma, N., Griffith, E.C., Harvey, C.D., Greenberg, M.E., 2021. Bidirectional perisomatic inhibitory plasticity of a Fos neuronal network. Nature 590, 115–121. https://doi.org/10.1038/s41586-020-3031-0

Yassin, L., Benedetti, B.L., Jouhanneau, J.-S., Wen, J.A., Poulet, J.F.A., Barth, A.L., 2010. An embedded subnetwork of highly active neurons in the neocortex. Neuron 68, 1043–1050. https://doi.org/10.1016/j.neuron.2010.11.029

Version 2

Abstract

Electroconvulsive therapy (ECT) is a highly effective treatment for several psychiatric disorders, though its biological mechanisms remain unclear. Its therapeutic action has traditionally been attributed to the generalized seizure ECT induces. However, this view is challenged by the recent finding that electroconvulsive stimulation (ECS) can trigger a cortical spreading depression (CSD). Because CSD triggers massive intracellular molecular changes, we hypothesized that it could be a key mediator of ECT’s therapeutic, plasticity-inducing effects. We observed similar neuronal oscillations following ECS in mice and patients undergoing ECT. We show that CSD drives increased expression of the immediate early gene Fos, a key marker of neuronal plasticity, and is associated with factors that predict positive ECT therapeutic outcome. Our results suggest that the therapeutic efficacy of ECT may be mediated by CSD. This challenges the seizure-centric model and implies that CSD, a currently unmonitored neurophysiological event, may serve as a more relevant biomarker for predicting and optimizing therapeutic outcomes of ECT.

Introduction

Electroconvulsive therapy (ECT) is one of the most potent and rapidly acting interventions in clinical psychiatry. It demonstrates substantial efficacy in treating severe and often treatment-resistant conditions, including major depressive disorder (Trifu et al., 2021), bipolar disorders (Tor et al., 2021), catatonia (Perugi et al., 2017), mania (Martin et al., 2025) and schizophrenia (Grover et al., 2019). ECT frequently achieves positive therapeutic outcomes in terms of remission rates and speed of response that surpass standard pharmacotherapies, making it indispensable in urgent clinical situations such as high suicide risk or severe catatonia (Baghai and Möller, 2008). ECT is also one of the few treatment options available to drug-resistant patients (Haddad and Correll, 2018). Despite its established clinical utility and decades of refinement aimed at enhancing safety and tolerability, the fundamental neurobiological mechanisms driving ECT’s therapeutic effects remain poorly understood.

There is a multitude of biomarkers affected by ECT that could underlie its mode of action as a psychiatric treatment (Singh and Kar, 2017). These include, non-exhaustively, the modulation of various neurotransmitter systems (Baldinger et al., 2014), alterations in regional cerebral blood flow (Takano et al., 2011) and metabolism (McCormick et al., 2007), changes in functional brain network connectivity (Ulrich et al., 2025), and the promotion of neuroplasticity (Bouckaert et al., 2014; Chen et al., 2018; Joshi et al., 2016). Understanding how these effects contribute to therapeutic outcomes is a significant challenge. We suspect that a critical gap in our knowledge lies in the early cascade of events triggered by an ECT treatment.

ECT, as it is used today, is targeted at the induction of a controlled seizure under general anesthesia and muscle relaxation through an electroconvulsive stimulation (ECS). Import to note here is that the term seizure likely refers to a different neuronal event in the context of ECT and epilepsy. In modern practice, ECS triggered seizures are highly synchronized brain-wide oscillations, which typically follow a three-phase progression (Brumback and Staton, 1982) consisting of an immediate 14-22 Hz activity (phase I), which evolves into arhythmic poly-spike patterns (phase II) and is then rapidly, within seconds, followed by a sustained low-frequency oscillation at 2-4 Hz (phase III). Phase III oscillations often occur without any ictal spikes that are characteristic of epileptic seizures (Brumback and Staton, 1982) and share limited similarities only with the late phase of seizures observed in epilepsy (Pottkämper et al., 2021). The phase III oscillation is referred to as seizure in the ECT literature. Illustrating the difference between what is referred to as a seizure in the ECT literature and what is referred to as a seizure in the epilepsy literature is also the fact that the post-seizure neuronal silencing of activity is a marker of treatment success in ECS, while it is associated with sudden unexpected death in epilepsy (Krystal et al., 1995; Sowers et al., 2013), and the finding that antiepileptics do not block ECS-triggered seizures (Cinderella et al., 2022). To try to disambiguate this, we will refer to the ECS induced seizure as phase III oscillation in the description of our results, but consistent with previous usage as seizure when talking about prior ECT work.

Historically, the generalized ECS induced seizure was considered to be the essential therapeutic component of the treatment (Accornero, 1988; Ulett et al., 1956). However, the occurrence and properties of seizures are only partially predictive of therapeutic outcomes (Perera et al., 2004; Sackeim et al., 1991). Similarly, the clinical efficacy of alternative stimulation methods, such as low amplitude ECT or magnetic seizure therapy, demonstrates that robust outcomes can be achieved with lower field strengths (Peterchev et al., 2015). This discrepancy suggests that electric field strength is unlikely to be the primary determinant of therapeutic success (Deng et al., 2024), indicating that ECT’s efficacy hinges on biological mechanisms beyond the mere presence of a seizure or the magnitude of the electrical stimulus.

A secondary neurophysiological event frequently triggered by seizures is cortical spreading depression (CSD). It has been speculated that CSD is an endogenous mechanism of the brain used for seizure termination (Dreier et al., 2012; Kelly et al., 1999). A CSD can also be triggered by direct electrical stimulation of the cortex (Fregni et al., 2007; Leao, 1944), and, consistent with this, recent evidence revealed the existence of CSD-like calcium events following ECS in mice and likely following ECT in patients (Rosenthal et al., 2025). This discovery necessitates a critical reevaluation of the traditional seizure-centric view of the effects of ECT. The profound neurophysiological disruption inherent to CSD may potentially act as a trigger for several of the neurobiological alterations driven by ECT.

CSD is characterized as a slowly propagating wave (typically 2–5 mm/min) of near-complete, and sustained depolarization affecting both neurons and glial cells (Seidel et al., 2016). This wave of intense depolarization is invariably followed by a period of marked suppression of spontaneous and evoked electrophysiological activity that lasts several minutes and can persist for up to an hour (Sawant-Pokam et al., 2017). The wavefront of the CSD is a massive transmembrane ionic flux (Somjen, 2001) that is accompanied by cellular swelling due to water movement, blood flow changes resulting in local hypoxia, and structural changes to dendritic processes (Takano et al., 2007). A CSD can be triggered by a large efflux of potassium ions into the extracellular space (Brinley et al., 1960) that causes a positive feedback loop via glutamate release, which propagates the wavefront through the activation of NMDA receptors (Zhou et al., 2013). Concurrently, there is an influx of sodium, chloride and calcium ions into cells (Pietrobon and Moskowitz, 2014). Disturbed calcium signaling is associated with a host of neuropsychiatric disorders treated by ECT (Nanou and Catterall, 2018), which makes CSD driven calcium influxes a likely key contributor to neuronal plasticity effects driven by ECT (Dehbandi et al., 2008; Wernsmann et al., 2006).

Here, we sought to characterize the specific calcium dynamics during ECS-induced CSD, hypothesizing they might underlie the plasticity-relevant effect of ECT. Consistent with previous work (Rosenthal et al., 2025), we observed a slow travelling calcium event that appeared to be a CSD. We found that the characteristics of this calcium event were consistent with being a CSD, and that it accounted for the ECS driven Fos expression, a marker of neuronal plasticity in the cortex (Bolhuis et al., 2001; Fleischmann et al., 2003; Mahringer et al., 2022, 2019). Furthermore, we show that EEG metrics known to predict positive therapeutic outcomes of ECT are correlated with the occurrence of CSD in mice. Our findings support the hypothesis that CSD, rather than the seizure itself, is the primary driver of ECS driven plasticity and, by extension, the therapeutic benefits of ECT.

Results

ECS triggers similar neuronal oscillations in both mice and humans

We first asked how well ECS in mice recapitulates EEG signatures of ECS in humans. To do this, we measured neuronal responses following ECS in mouse dorsal cortex, using concurrent widefield calcium imaging and EEG recordings (Methods). We expressed a genetically encoded calcium indicator brain-wide by retroorbital deposit of an AAV-PHP.eB-Ef1α-GCaMP6s-WPRE and performed concurrent EEG recordings using a semitransparent 30-channel EEG electrode (Figure 1A). During ECS, mice were anesthetized with isoflurane (2%) and myorelaxed with succinylcholine chloride (1mg/kg). ECS was delivered through biauricular electrodes using a commercial ECS unit. Stimulation parameters were set within a range that reliably evoked tonic-clonic seizures in mice (Methods).

Electroconvulsive stimulation (ECS) triggers similar slow neuronal oscillations in mice and humans.

(A) Top: Schematic of mouse ECS setup. We performed concurrent widefield calcium imaging and EEG in the mouse dorsal cortex during biauricular ECS. Bottom: Layout of the EEG probe. (B) Spectrograms of two widefield calcium imaging ROIs surrounding EEG electrodes L12 and R12 (as highlighted in A). Duration of the ECS is marked by a black bar. (C) Spectrograms of the EEG recorded concurrently with the widefield calcium data in C. Time during which EEG is not recorded due to a stimulation artifact is marked by gray shading. (D) Schematic of patient ECS setup. EEG recordings were performed using two frontal electrodes after bifrontal ECS in patients. (E) Spectrograms of EEG recorded following ECT in a patient. Patient EEG recordings are started following ECS and terminated when the oscillation stops. Gray shading marks times without EEG recording. (F) Top: Oscillations observed in early, mid, and late phases after ECS in an example mouse widefield calcium ROI (blue) and in the concurrently recorded EEG (green). Bottom: Oscillations observed in an example patient EEG recording. (G) Oscillation duration for mouse and patient EEG recordings. The central box represents the interquartile range (IQR), spanning from the first (Q1) to the third (Q3) quartiles. Whiskers extend to data points not considered outliers (Methods). A central horizontal line indicates the median and an open circle the mean of the data distribution. (H) Power-weighted mean frequency at peak and offset of oscillations for mouse widefield, mouse EEG and patient EEG recordings. Here and elsewhere: n.s.: not significant, *: p < 0.05, **: p < 0.01, ***: p < 0.001. See Table S1 for all information on statistical testing.

Across the parameter range used, ECS triggered an immediate and brief calcium response across the entire dorsal cortex (Figure S1A). The response onset occurred with a latency of less than 10 ms (within the first frame after stimulation at 100 Hz acquisition), and calcium activity continued to increase until the end of the ECS (Figure S1B). In the EEG data (Figure S1C) we were blind to this direct response, as we disconnected the EEG from the data acquisition system during ECS to prevent electrical arcing between electrodes (Methods). We speculate that the direct ECS response observed in calcium imaging is driven by spiking activity in the cortical neurons, given that the electrical field generated by the ECS is several times higher than the neuronal activation threshold (Deng et al., 2011). Following the direct calcium response to ECS, mean calcium activity decreased and returned to baseline only slowly (Figure S1B).

Following ECT, the combined median duration of phases I and II oscillations is 9 s in patients (Hogan et al., 2019). Consistent with this and previous reports (Murakami et al., 2008), we observed a phase III slow oscillation in the delta to low theta range (1 to 5 Hz) in mice that emerged about 10 s after the ECS, both in calcium activity and EEG recordings (Figure 1B-C, S2). This oscillation was strikingly similar to that observed in patients undergoing ECT (Figure 1D-E) (Stuiver et al., 2026), and was present in the EEG recording of all patients analyzed here. Given that ECT stimulation parameters are chosen to increase the likelihood of phase III oscillations, which in turn are used as a biomarker for treatment outcome (Sackeim et al., 1991), this is perhaps not surprising.

The oscillation followed a similar pattern in both mice and patients and typically exhibited a strong sinusoidal shape in the early stage that desynchronized over time (Figure 1F). In mice, the oscillations lasted 42.3 ± 11.9 s (mean ± std), and we found no evidence of a difference in oscillation duration between mice and patients (Figure 1G). These values are consistent with previous estimates of the duration of phase III oscillations in patients (Hogan et al., 2019). Interestingly, the frequencies of the oscillations were also very similar in mice and patient EEG, and both tended to decrease with time (Figure 1H). The estimate of the oscillation frequency in the mouse calcium imaging data was systematically higher relative to the concurrent EEG data (Figure 1H) but exhibited the same decreasing trend. We suspect this difference is driven by differential contributions to the two signals from non-oscillation related neuronal activity and noise sources.

ECS can trigger a travelling calcium event

We found that ECS could also trigger a travelling calcium event, as previously described (Rosenthal et al., 2025). It has been speculated that this calcium event is a CSD (Rosenthal et al., 2025), and while we agree with this interpretation, we will refer to it as a calcium event in the results section here to avoid forgoing the conclusion. To monitor the propagation of these events, we recorded the calcium activity across the entire dorsal cortex for 3 to 5 minutes following ECS (Figure 2A). They manifested as a significant increase in calcium fluorescence that propagated across the dorsal cortex (Figure 2B) at a speed of 0.083 ± 0.009 mm/s (mean ± std) (Figure 2C), consistent with previously reported CSD propagation speeds in mice (Enger et al., 2015). Calcium events exhibited maximum fluorescence intensities that exceeded even those observed during the direct ECS response (Figure 2D). Once initiated, these events spread across an entire dorsal cortex hemisphere over the course of several minutes (Figure 2E). In line with the refractory dynamics of CSD, the event exhibited a non-repeating trajectory, traversing each part of dorsal cortex only once. The lateralization of the events exhibited a certain stochasticity and calcium events were not always symmetrically present in both hemispheres. We frequently observed cases in which the events were unilateral and constrained to only left or right cortical hemispheres (Figure 2F-G).

ECS can trigger a travelling calcium event.

(A) Left: Schematic of mouse ECS setup without concurrent EEG recordings. Right: Widefield recordings of mouse dorsal cortex were parcellated into 818 ROIs, each of which was assigned to a cortical region according to the Allen Brain Atlas taxonomy. The ROIs highlighted correspond to the calcium activity shown in B and E. (B) Example of calcium activity in ROIs highlighted in A during the traversal of a calcium event. (C) Propagation speed of calcium events (Methods). Each data point is the median speed across one calcium event. (D) Maximum calcium activity throughout the dorsal cortex during pre-ECS, direct ECS response and calcium event periods. Two direct ECS responses are outside the plot range and not shown (1707 % and 2374 % ΔF/F0). (E) Left: Low-passed (< 0.5 Hz) calcium activity during an example ECS session triggering calcium event over both hemispheres. Each panel shows the average calcium activity over a 10 ms window starting at the time indicated above the panel. Right: Isochrones (spaced at two seconds increments) indicating the time at which each region reaches 200% of its baseline fluorescence. (F) As in E, but for a unilateral calcium event over the right hemisphere following a 150 s delay post-ECS. (G) As in E, but for a unilateral calcium event over the left hemisphere.

In about one third of cases, calcium events appeared bilaterally, while in the remaining two thirds the events were confined to one hemisphere only (Figure 3A, Video S1). The majority of unilateral events occurred in the right hemisphere, and across all events, most first appeared on the right edge of the cranial window (Figure 3B). This right hemisphere bias was likely driven by the ECS polarity, as we always placed the positive pole of stimulation on the right ear (Figure 1A). Calcium events were the only type of response to ECS that occurred unilaterally (Figure 3C). While all calcium events co-occurred with oscillations, just over half of the oscillations were accompanied by a calcium event (Figure 3D). The likelihood of triggering a calcium event has been shown to be a function of the stimulation current and frequency used of the ECS (Rosenthal et al., 2025). To investigate how the stimulation parameters control the appearance of oscillations and calcium events, we tested stimulation frequencies between 25 and 100 Hz and stimulation currents between 15 and 75 mA for a total charge delivery of between 0.187 and 3.750 mC. With most combinations of parameters, we observed a non-zero likelihood of triggering a phase III oscillation (Figure 3E). The threshold for triggering calcium events was systematically higher and appeared to be a function of both stimulation frequency and current (Figure 3E). Based on this, we hypothesized that total charge delivery might be a good predictor of the occurrence of calcium events. A linear regression of total stimulation charge against the likelihood of triggering a calcium event showed a significant correlation (Figure 3F). Stimulation frequency and stimulation current also exhibited a positive but nonsignificant correlation with the likelihood of triggering a calcium event (Figure S3A). A multivariate linear regression confirmed that the interaction term of current and frequency was significant and non-zero (0.015 % / (mA Hz)). We found a similar relationship between the total stimulation charge and the probability of triggering a phase III oscillation (Figure 3G, S3B). Consistent with the hypothesis that the calcium event exhibits all-or-none dynamics, we found no evidence of changes in propagation speed, total duration, or delay to onset of calcium events as a function of total stimulation charge (Figure S3C-E). We also observed that the total ECS charge correlated with the intensity of the direct ECS response (Figure S4A). This direct response, in turn, predicted the hemisphere of the calcium event in 75% of the unilateral cases (Figure S4B), supporting our hypothesis that the ECS stimulation parameters can control the occurrence of calcium events.

Calcium events are triggered preferentially by higher total ECS charge.

(A) Probability distribution of the laterality of calcium events. (B) Spatial distribution of calcium events detection sites. For bilateral calcium events, only the hemisphere where the calcium event was detected earliest is considered as the detection site. (C) Fraction of stimulation responses, oscillations and calcium events that were bilateral. (D) Probability of oscillations being observed if a calcium event is triggered (P(osc. | calcium event)), of a calcium event being observed if an oscillation is triggered (P(calcium event | osc.)), and of neither being observed (P(neither)). (E) ECS parameter map of the likelihood of triggering a calcium event (top) or an oscillation (bottom). (F) Probability of triggering a calcium event as a function of total ECS stimulation charge. The solid line represents the linear fit of the data; the shaded region represents 95% confidence interval on the fit. (G) As in E, but for the probability of triggering an oscillation.

The calcium event exhibits hallmark features of CSD

The mode of propagation, speed and fluorescence intensity support the hypothesis that the travelling calcium event is a CSD. This has already been speculated previously (Rosenthal et al., 2025), and makes a number of directly testable predictions. First, a CSD changes the local ionic homeostasis and is accompanied by hemodynamic constriction (Takano et al., 2011). To test for hemodynamic effects triggered by ECS, we performed widefield imaging experiments in a cohort of mice in which we expressed enhanced green fluorescent protein using an AAV (EGFP,AAV-PHP.eB-Ef1α-EGFP-WPRE). In widefield imaging, the EGFP fluorescence signal depends on pH and overlying blood volume. The latter is a phenomenon called hemodynamic occlusion, whereby decreased blood vessel diameter increases the EGFP signal (Ma et al., 2016; Yogesh et al., 2025). During the ECS, we observed an immediate increase in the EGFP signal that lasted for a couple of seconds and is consistent with acute vasoconstriction (Figure S5A). We also observed travelling signals in the EGFP recordings with spatial and temporal characteristics that resembled the calcium event (Figure S5B). The amplitude of these events was smaller than that observed in calcium imaging (Figure S5C), but they propagated at a speed that did not differ significantly from that observed in calcium events (0.096 ± 0.02 mm/s, Figure S5D). This would be consistent with either vasoconstriction or an increase in intracellular pH. CSD is associated with vasoconstriction (Takano et al., 2007), but typically only a brief alkalization over the course of seconds followed by a more prolonged acidification of the intracellular domain over the course of minutes (Sun et al., 2011). However, given that an estimated alkalization of less than 0.5 in pH, would translate to less than a 10% ΔF/F0 change in EGFP fluorescence (Kneen et al., 1998), we suspect most of the change we observe is driven by hemodynamics.

A second feature of CSD is the propagation through extracellular diffusion (Enger et al., 2015), rather than through electrical activity. Increases in intracellular calcium levels in neurons are typically associated with increases in spiking activity. A CSD, however, is associated only with brief increases in spiking activity at the onset (Herreras et al., 1994; Nasretdinov et al., 2023). To visualize the propagation of the calcium event in the tissue, we used two-photon microscopy to observe the wavefront at cellular resolution. We expressed a genetically encoded calcium indicator (AAV2/9-Ef1α-GCaMP6f-WPRE) in visual cortex and measured responses to ECS in layer 2/3 neurons (Figure 4A, Video S2). Consistent with a diffusion-limited process, we found that the calcium event was a sharp wavefront of significant increase in fluorescence, propagating across the entire field of view (Figure 4B) at a speed of 0.079 ± 0.0174 mm/s (mean ± std). This aligns with our speed estimates based on widefield calcium imaging. Combined with the lack of propagation along inter-hemispheric or other known anatomical projections (Figure 2E-G), this supports the idea that the calcium events are propagating through a diffusion-limited process, rather than synaptic activity, consistent with a CSD.

Calcium events first spread in supragranular layers and elicit long-lasting calcium increase in cells.

(A) Schematic of the experimental setup. We performed two-photon (2p) calcium imaging in the right primary visual cortex during ECS. (B) 2p calcium activity during an example ECS session triggering a calcium event. Time of maximum (TOM) is shown on the right. Each panel shows the average calcium fluorescence over a 500 ms window starting at the time indicated above the panel. Gray inset highlights two example somata whose calcium signals are shown in G. (C) Top: Schematic of the micro-prism implanted below the imaging window to allow for simultaneous recording in all cortical layers (see Methods). Bottom: Widefield imaging of the implant 18 seconds after ECS, confirming the prism does not obstruct the calcium event propagation. A gray arrow indicates the orientation of the prism. (D) As in B, but imaged through a prism implant. Cortical layers 3, 4, and 5a (L3, L4, L5a) are defined based on soma densities (Methods). (E) Calcium fluorescence of superficial (blue) and deep (green) layers somata during calcium events, aligned to the time of maximum of all neuropil (Methods). Lines represent the hierarchical bootstrap estimates of the mean value for each time bin. Here and elsewhere, shading around the mean is one standard deviation of the bootstrap distribution at each time bin. Response curves are compared for each time bin: in the horizontal bars above the plot, black marks time bins where p<0.05 and gray mark time bins where p>0.05. (F) Time of 50% of maximum activity for superficial and deep layers somata during calcium events. (G) Top: Calcium fluorescence from the two example somata in B. Bottom: Calcium fluorescence from the neuropil surrounding these somata. Fitted slopes during the 60 seconds following calcium event onset are shown as dashed lines. (H) Calcium fluorescence of somata (dark gray) and neuropil (purple) in all layers during calcium events. Each soma is aligned to the time of maximum of the neuropil. (I) Distribution of the fitted slopes for cells and neuropil during the 60 seconds following calcium events. (J) Scatter plot of fitted slopes and correlation of cells with their neuropil. Top and right histograms are marginal distributions over slopes and correlations, respectively.

A third hallmark feature of CSD is its preferential propagation through superficial cortical layers (Stafstrom, 2020; Zakharov et al., 2019). To investigate this, we implanted a micro-prism in the cortex, enabling simultaneous imaging across all cortical layers (Figure 4C, Video S3). We found that the calcium event first spread in superficial layers 2/3 and 4, and only later reached deeper layers 5 and 6 (Figure 4D). Calcium event response in superficial layers preceded those in deep layers by 6.9 ± 1.3 s (mean ± std) (Figure 4E-F). While both somata and neuropil showed a sharp calcium fluorescence increase at the wave front, calcium decay kinetics were more heterogeneous. Calcium fluorescence in the neuropil gradually returned to baseline over the course of about 100 s. Most somata followed this time course, but a subset of them exhibited a progressive increase in fluorescence (Figure 4G), likely reflecting a slow accumulation of somatic calcium. This resulted in a population average fluorescence decrease that was significantly faster in the neuropil than in the somata (Figure 4H-I). The somata with increasing post-calcium event fluorescence made up about 35% of the total somatic population (Figure 4J). This heterogeneous response pattern is a characteristic of CSD, and is hypothesized to be linked to the variable capacity of individual neurons to recover from CSD-induced oxidative stress (Enger et al., 2015). Overall, these results strongly support the hypothesis that the ECS triggered calcium events are indeed CSD.

Calcium events, not ECS, drives Fos expression

If those calcium events are CSD, then they should cause neuronal plasticity (Faraguna et al., 2010; Li et al., 2012). To begin testing this hypothesis, we quantified the ECS driven expression of the Fos protein, the product of an immediate early gene that is a marker of functional plasticity in cortex and hippocampus (Mahringer et al., 2022, 2019). We again performed widefield calcium imaging of ECS responses and allowed one hour of recovery post-ECS for Fos expression (Methods). Brain tissue was then processed for Fos immunohistochemistry. We found that ECS drove strong Fos expression in a variety of brain areas, including the cortex (Figure 5A). To determine whether the direct electrical response to the ECS or the subsequent calcium event drove cortical Fos expression, we again took advantage of the occurrence of unilateral calcium events (Figure 5B). This allowed us to directly compare Fos expression between hemisphere that exhibited a calcium event and those that did not. Fos was also upregulated in subcortical structures (e.g. amygdala, ventromedial hypothalamus) and in cortical regions outside our widefield imaging field of view (e.g. dentate gyrus, piriform cortex); however, we remain agnostic as to whether the CSD invaded these areas. Notably, within our recorded cortical regions, hemispheres without a calcium event exhibited Fos levels that were statistically indistinguishable from sham-treated controls. Instead, the ECS-driven increase in Fos expression was fully accounted for by the occurrence of a calcium event (Figure 5C). This supports the hypothesis that calcium events, which are likely CSD, are the primary driver of neuronal plasticity following ECS.

The ECS triggered increase in Fos expression in cortex is driven by the calcium event.

(A) Example of coronal brain sections stained for Fos from three sham ECS (top) and three ECS-treated (bottom) mice. (B) Maximum calcium fluorescence during direct ECS response (first row), the amplitude of the ECS induced oscillation (second row), and the peak fluorescence to illustrate the presence or absence of a calcium event (third row). From left to right, data from three example mice that had unilateral left, unilateral right, and bilateral calcium events, respectively. The bottom row shows corresponding coronal sections from the same brains, stained for Fos expression. (C) Quantification of the Fos throughout the entire cortex, expressed as a fold change for an entire hemisphere normalized to sham mice (see Methods). Paired hemispheres (experiments in which ECS triggered a bilateral direct ECS response and oscillation followed by a unilateral calcium event) are linked by a line. Thus, the calcium event, not the direct ECS response or the phase III oscillation, is the correlate of Fos expression.

The occurrence of calcium events correlates with EEG metrics predicting positive ECT therapeutic outcomes

Given that calcium events accounted entirely for the expression of Fos, we speculated that they may also be the driver of the positive therapeutic outcomes of ECT. A first testable prediction of this hypothesis is that EEG responses to ECS that correlate with positive therapeutic outcomes should also correlate with the occurrence of calcium events. While EEG responses are only partially predictive of clinical improvement following ECT (Krystal and Weiner, 1994), there are several EEG metrics that correlate with positive therapeutic outcomes. For the phase III oscillation, an increased amplitude (Perera et al., 2004), a longer duration (Gillving et al., 2024), a higher post oscillation suppression index (Kimball et al., 2009; Nobler et al., 1993), and higher interhemispheric coherence (Grözinger et al., 2013), for example, are indicators of a favorable patient response to ECT (Francis-Taylor et al., 2020).

To address this, we investigated whether these EEG metrics correlated with the occurrence of a calcium event, by comparing EEG data from recordings without and with calcium events (Figure 6A-B). We found that the mid-phase III oscillation amplitude was significantly higher in hemispheres where a calcium event occurred compared to the unaffected contralateral hemisphere or to ECS where neither hemisphere showed a calcium event (Figure 6C). Both the oscillation duration and postictal suppression index (Figure 6D-E) were increased in ECS triggering a calcium event, but we found no evidence that the laterality of the calcium event (ipsi-vs. contralateral) influenced these two metrics. This absence of hemisphere specificity would be consistent with the observation of a tight coupling of onsets and offsets of phase III oscillations across the hemispheres (Brumback and Staton, 1982). In the case of inter-hemispheric coherence of the EEG signals, we found no evidence of an influence of the calcium event (Figure 6F). These results led us to speculate that calcium events may serve as a central mechanistic driver of the therapeutic efficacy of ECT.

Calcium events are associated with ictal EEG metrics that predict clinical ECT efficacy.

(A) Example of an ECS session triggering a direct ECS response followed by a slow oscillation. Data is shown for two example electrodes. A gray transparent square indicates the duration of the oscillation. First row: Calcium imaging spectrogram of the ROI surrounding the electrode. Second row: EEG signal spectrogram of the electrode. Third row: z-scored EEG signal in the 1-5 Hz band to visualize the oscillation amplitude. Fourth row: EEG Hilbert amplitude in the 1-30Hz band, with dashed lines indicating the mean amplitudes during and after the oscillations used to compute the postictal suppression index (PSI). Fifth row: Calcium imaging fluorescence, low-passed below 0.5 Hz. (B) As in A, with identical ECS stimulation parameters, but for an ECS session triggering a direct ECS response followed by a bilateral calcium event. (C) Mid-oscillation amplitude for each electrode during ECS sessions triggering only direct ECS response (gray), ECS sessions triggering a unilateral calcium event over the contralateral hemisphere (blue) and ECS sessions triggering a calcium event on the ipsilateral hemisphere (red, either unilateral or bilateral calcium event). See Methods for the details on the ictal EEG metrics definitions. (D) As in C, but for the oscillation duration. (E) As in C, but for the postictal suppression index. (F) As in C, but for the inter-hemispheric coherence of the EEG signal.

Discussion

Here, we sought to characterize the neuronal response to ECS, focusing on a calcium event that shows hallmark features of a CSD. We demonstrate that these events correlate with prognostic EEG markers of ECT treatment outcomes and induce the expression of Fos, a key indicator of neuronal plasticity. Based on our results, we propose that CSD may be a more relevant biomarker than the phase III oscillation to explain the therapeutic effects of ECT.

Limitations

There are two key limitations that constrain the interpretation of our results, and that will need to be addressed in future research:

We have no direct evidence that patients also have ECT-induced CSD. Patients are typically monitored with two-channel EEG to determine whether ECT has induced a phase III oscillation (Francis-Taylor et al., 2020), but these measurements are currently not sufficient to detect CSD. One simple way to modify the current clinical routine measurements would be to perform direct current (DC) EEG recordings with a higher channel count. CSD results in characteristic DC shifts in the EEG response (Hartings, 2016). While detecting stochastically-occurring CSD signatures in EEG recordings of awake subjects is difficult (Lauritzen et al., 2011), this would likely be facilitated in the context of ECT by the known timing of the triggering stimulus and the elimination of movement artifacts via anesthesia and myorelaxation (Hofmeijer et al., 2018). Alternatively, given that the CSD triggers a strong hemodynamic change (Figure S5) (Hadjikhani et al., 2001), blood-oxygenation level dependent imaging techniques might be able to detect a CSD. There is already evidence of hemodynamic changes induced by ECT in patients (Rosenthal et al., 2025).

Our conclusion that the CSD is the primary driver of plasticity is based on its role in driving Fos expression, an early marker of cortical plasticity (Mahringer et al., 2022, 2019). We cannot rule out the possibility that the seizure induced by ECS drives Fos independent plasticity pathways that also contribute to therapeutic outcomes. Like CSD, seizures are neuronal events characterized by supra-physiological excitability that trigger neuronal plasticity (Scharfman, 2002). While knocking out Fos reduces seizure-induced plasticity (Watanabe et al., 1996), other signaling cascades such as mTOR (Kim et al., 2022), BDNF (Ramnauth et al., 2022) and the immediate early gene Arc (Larsen et al., 2005) can all be triggered by ECS-induced seizures. One way to explore the contribution of these plasticity pathways to treatment effects will be to map the gene expression changes following ECT more broadly.

Does ECT drive CSD in humans?

In addition to the results presented in this paper and that of (Rosenthal et al., 2025), there are three additional lines of evidence that support the hypothesis that ECT triggers CSD in patients:

A significant proportion of ECT patients suffer from a migraine-like headache immediately after the treatment (Mulder and Grootens, 2020). Migraines are thought to be caused by a CSD driven calcium influx that activates Panx1 channels (Karatas et al., 2013).

Confusion and cognitive side effects, with a particular focus on memory and hippocampus-dependent processes, are frequently reported following ECT (Impastato and Almansi, 1942; Porter et al., 2020). In mouse models of epilepsy, postictal confusion has been shown to be driven by hippocampal CSD, rather than by the seizure itself (Mitlasóczki et al., 2025).

CSD causes osmotic changes (Mazel et al., 2002) that could explain the structural imaging changes observed after a single ECT session (Takamiya et al., 2025; Verdijk et al., 2026), which are particularly pronounced in the hippocampus (Bracht et al., 2026; Wilkinson et al., 2017).

Our data show that metrics which predict a positive ECT treatment outcome also correlate with the occurrence of CSD (Figure 6). Given that standard ECT protocols optimize for these metrics as a proxy for seizure quality (Krystal et al., 1995; Sackeim et al., 1991), it is possible that clinical practice has inadvertently optimized ECT to trigger CSD. If CSD accounts for plasticity effects, triggering a CSD in a non-seizure context may be sufficient to elicit therapeutic effects. This is supported by the use of ultrasound stimulation which functions as a treatment for major depressive disorder (Oh et al., 2024) and can trigger CSD in mice (Estrada et al., 2021). Much like using a landslide to create ripples in a pond, a generalized seizure may be an unnecessarily burdensome way to trigger the CSD required for treatment-relevant plasticity effects. Conversely, the accumulation of calcium transients over prolonged periods of time may be sufficient to drive immediate early gene expression and trigger treatment-relevant plasticity without CSD. This is supported by the cell-type specific increase of small calcium transients during rTMS (Gongwer et al., 2025), which may activate signaling pathways triggered by cumulative calcium load that result in downstream immediate early gene expression, such as the CaMKIV-CREB pathway (Gandolfi et al., 2017; Wu et al., 2001). This makes a testable hypothesis that the effect of rTMS depends on the total calcium influx accumulated over the course of a treatment, consistent with the greater efficacy of higher-dosing stimulation protocols (Cole et al., 2022).

CSD as a driver of ECT related plasticity

The massive intracellular calcium influx during CSD (Figure 2, 4) is a prime candidate for driving the molecular changes underlying ECT-induced neuronal plasticity. Calcium is the entry point for a variety of plasticity-related signaling pathways that are disrupted in psychiatric conditions (Nanou and Catterall, 2018). While both physiological activity and seizures can trigger these pathways, CSD stands out due to the sheer magnitude of its calcium surge, driven in part by a near-total release of endoplasmic reticulum stores (Stern et al., 2024). Elevated calcium levels drive the expression of key plasticity markers. As such, CSD is known to induce Fos in non-ECS contexts (Ingvardsen et al., 1997; Iqbal Chowdhury et al., 2003) and causes differential expression of a host of plasticity-related genes (Dell’Orco et al., 2023). One of the many downstream effectors of these changes is increased BDNF expression (Faraguna et al., 2010). As the primary ligand for the TRKB receptor, BDNF activates a pathway that is a well-established target for both antidepressant (Casarotto et al., 2021) and antipsychotic drugs (Pandya et al., 2013).

Our observation that ECS drives Fos expression in the cortex following CSD contrasts with the reports that ECS fails to increase, or even downregulates, cortical Fos expression (Calais et al., 2013; Winston et al., 1990). There are two possible explanations for this. Because CSD occurs stochastically (Figure 3), experiments that do not monitor for it cannot attribute the presence or absence of Fos to ECS itself – a ECS without CSD results in cortical Fos levels not significantly different from sham ECS (Figure 6). Alternatively, repeated ECS may engage a meta-plasticity mechanism, in which successive inductions progressively dampen Fos expression, for instance through histone acetylation (Park et al., 2014).

While we cannot fully exclude the possibility of a secondary event that could modulate Fos expression, such as a unilateral seizure restricted to the CSD hemisphere, we monitored calcium activity in a subset of mice (11/17) until shortly before euthanasia and did not see any seizure-like calcium activity. We also did not observe behavioral signatures of a seizure in any of the mice. But even if secondary events did occur, the first unilateral event we observed was the CSD. This, combined with the fact that the CSD provides a potent biological mechanism to explain the robust neuroplasticity driven by ECT, leads us to hypothesize that the CSD may be the primary treatment relevant event triggered by ECT, and not the phase III oscillation.

Origin of phase III oscillation

In the ECT literature, the phase III oscillation is often referred to as a seizure. We suspect that the key difference to an epileptic seizure is that an initial strong input (either directly from the ECS or indirectly via ECS driven excitation of cortex) to thalamus triggers a 2-4 Hz oscillation (Bal et al., 2000). At the same time, ECS stochastically triggers CSDs. As these move across cortex they silence cortical activity, and what remains in their wake is the oscillatory drive from thalamus. A central thalamic is consistent with the observation that phase III oscillations have zero phase lag across the human cortex (Brumback and Staton, 1982; Huels et al., 2023). In response to ECS, this gives rise to a phase III oscillation that does not have the ictal spikes characteristic of epileptic seizures. Ictal spikes in epileptic seizures are thought to reflect cortical responses to the oscillating thalamic drive (Meeren et al., 2002; Polack et al., 2009). Interestingly, a combination of CSD induced plasticity and the phase III oscillation (Figure 3D) could result in a selective strengthening of thalamocortical synapses. It has been speculated that such an increase is the mechanism that underlies the therapeutic efficacy of ECT (Wei et al., 2020). Reduced coupling between thalamus and frontal areas of cortex is characteristic of schizophrenia (Giraldo-Chica and Woodward, 2017; Vinogradov et al., 2023), and assuming this decoupling is directly related to the etiology of the disease (Keller and Sterzer, 2024), this mechanism of action could explain both the efficacy of ECT in treating psychosis and why patients with psychotic depression have more favorable ECT treatment outcomes than those with nonpsychotic depression (Petrides et al., 2001).

Future directions

Unresolved is the question of how to reconcile the non-specific nature of CSD with the cell type specific effects that characterize rTMS (Gongwer et al., 2025), antidepressant (Shao et al., 2025; Vesuna et al., 2020) and antipsychotic drugs (Heindorf and Keller, 2024; Lupori et al., 2026). Several lines of evidence suggest that the effects of CSD may also be cell type specific. First, CSD is easier to trigger in superficial than in deeper layers of cortex (Leo and Morison, 1945), and affects superficial somata for a significantly longer duration (Figure 4). This aligns with the evidence that schizophrenia is associated to a number of plasticity-related genes whose expression is selectively enriched in upper cortical layers (Batiuk et al., 2022). Second, the large spectrum of time courses of calcium responses to CSD (Figure 4) could be explained by cell type intrinsic mechanisms, such as differential NMDA receptor expression (Buchanan et al., 2012). Testing this hypothesis will require cell type specific chronic calcium imaging following ECS.

The immediate clinical challenge stemming from our findings will be to determine whether the role of CSD is therapeutically beneficial or potentially deleterious. Optimizing ECT involves balancing the likelihood of positive therapeutic outcomes while minimizing adverse cognitive effects, most notably retrograde amnesia and impairments in autobiographic memory (Porter et al., 2020; Sackeim et al., 2007). For example, there is an optimal window for the stimulation charge: while a minimum stimulation charge is required for therapeutic benefit, a stimulation charge significantly exceeding the phase III oscillation threshold is linked to increased cognitive side effects (Delva et al., 2000; Sackeim et al., 1987). This could be explained if a higher stimulation charge is more likely to trigger CSDs in brain regions more distant from the stimulation electrodes. For example, it is conceivable that a CSD in frontal cortex is therapeutically beneficial while a CSD in hippocampus is responsible for retrograde amnesia. In mice, the ECS driven CSD likely invades hippocampus (Figure 5) (Mitlasóczki et al., 2025). If so, this might open up potential for pharmacological interventions to prevent a CSD invasion of hippocampus during ECT, e.g. with NMDA receptor antagonists that preferentially affect hippocampus (Mutel et al., 1998). By restricting CSD to the neocortex, it may be possible to prevent retrograde amnesia without compromising the cortical plasticity effects required for therapeutic outcomes.

Relevance

Our findings suggest that CSD, a currently unmonitored neuronal response to ECS, appears to be the primary driver of ECT-induced neuroplasticity. Ultimately, we propose that monitoring and facilitating CSD during ECT may unlock the possibility to transform psychiatry’s most potent intervention into a targeted and reliable therapy with minimal side effects.

Methods

Mice

A total of 66 mice (27 male, 39 female), 7 – 22 weeks old at the start of the experiment, were used. Most mice were C57BL/6 wild-type, except for transgenic animals used for two-photon prism imaging experiments. In this case, transgene expression was unrelated to the measurements reported here and did not affect experimental readouts. Mice used for two-photon prism imaging were PlexinD1-CreERT2 animals in which ChrimsonR expression had been induced in the retrosplenial cortex (via tamoxifen food). No optogenetic experiment was performed here, and this co-expression does not affect GCaMP stimulation. See Table S2 for details of mouse inclusion for the different experiments. Mice were group-housed in a vivarium (light/dark cycle: 12/12 hours), and experiments carried out in light cycle. In female mice, we did not time experiments to a specific phase of the estrous cycle. All animal procedures were approved by and carried out in accordance with the guidelines of the Veterinary Department of the Canton of Basel-Stadt, Switzerland.

Surgery

For all surgical procedures, mice were anesthetized with a mixture of fentanyl (0.05 mg/kg; Actavis), midazolam (5.0 mg/kg; Dormicum, Roche), and medetomidine (0.5 mg/kg; Domitor, Orion) injected intraperitoneally. Analgesics were applied perioperatively (2% lidocaine gel, meloxicam 5 mg/kg) and postoperatively (3.25 mg/kg; Ethiqa XR). Eyes were covered with ophthalmic gel (Virbac Schweiz AG). For widefield imaging, we implanted crystal skull (Kim et al., 2016) or thinned skull (Yang et al., 2010) cranial windows. Briefly, the perimeter of the skull plate overlying the cortex was thinned using a dental drill until it could be removed with forceps (crystal skull) or only thinned in its entirety until transparent but not removed (thinned skull). To protect the brain, a glass or acrylic polymer window was then implanted. For two-photon imaging, a cranial window was implanted over primary visual cortex (V1) as previously described (Keller et al., 2012; Leinweber et al., 2014). Briefly, a 4 mm craniotomy was made over the right V1, centered 2.5 mm lateral and 0.5 mm anterior to lambda. The exposed cortex was sealed with a 3 mm or 4 mm circular glass coverslip. For prism imaging, a 0.7 mm glass prism (Tower Optical Corporation) was attached to the coverslip with UV-curable glue (Norland Optical Adhesive), and a small incision of cortical tissue was made to allow for the insertion of the prism into the cortex. Cranial windows were sealed using gel superglue (Ultra Gel, Pattex), and, for widefield windows, cyanoacrylate adhesive (Vetbond, 3M) (Kauvar et al., 2020). The remaining exposed surface of the skull was covered with cyanoacrylate adhesive, and a titanium head bar was fixed to the skull using dental cement (Paladur, Heraeus Kulzer). After surgery, anesthesia was antagonized by a mixture of flumazenil (0.5 mg/kg; Anexate, Roche) and atipamezole (2.5 mg/kg; Antisedan, Orion Pharma) injected intraperitoneally.

Virus injections

For widefield experiments, we deposited an AAV vector with PHP.eB capsid (Chan et al., 2017) retro-orbitally (4 μl per eye of at least 1014 GC/ml) to drive expression throughout cortex of GCaMP6s (AAV-PHP.eB-Ef1α-GCaMP6s-WPRE) or EGFP (AAV-PHP.eB-Ef1α-EGFP-WPRE). For two-photon experiments, the AAV was injected locally in the primary visual cortex using a pneumatic injector (IM-300 Microinjector, Narishige) to drive expression of GCaMP6f (AAV2/9-Ef1α-GCaMP6f-WPRE, 100-250 nL of at least 1014 GC/ml). Imaging commenced at the earliest 3 weeks after AAV injection.

Electroconvulsive stimulation (ECS) in mice

For all ECS experiments mice were anesthetized with isoflurane (2%) using 95% O2 as a gas carrier (1 L/min). For myorelaxation we injected succinylcholine chloride (1 mg/kg; Sigma-Aldrich) intraperitoneally. Mice were head-fixed, and ECS was delivered through auricular electrodes (Ugo Basile model #57802) using a pulse generator (Ugo Basile model #57800-001). A conductive salt-free electrolyte gel (Spectra 360 gel, Parker Laboratories) was applied to the electrodes to improve conductivity. Stimulation consisted of a 1 s long pulse train with a frequency of between 25 and 100 Hz, and a current amplitude of between 15 and 75 mA. Pulse width was 0.5 ms, which resulted in a total charge range of between 0.187 and 3.750 mC. These parameters reliably evoke tonic-clonic convulsions in mice (Hagen et al., 2015; van Buel et al., 2017). The distribution of ECS parameters and the specific parameters used for each single ECS example can be found in Table S3.

For comparison, patient ECT parameters used here were typically as follows: train duration up to 8 seconds, pulse frequency of between 40-60 Hz, current amplitude of between 890-920 mA, pulse width of between 0.5-0.75 ms, charge maximum 1000 mC. Thus, stimulus duration, current amplitude and resulting total charge are adapted to the mouse, but frequency/pulse-width are within clinical range (Peterchev et al., 2010). The list of ECT parameters used here can be found in Table S4. The stimulation polarity was constant for mice ECS, whereas bidirectional pulse train (alternating polarity) was used for patient ECT. For sham ECS experiments, we used an identical procedure with no current delivery. Up to 6 ECS sessions were administered at intervals of 48 hours. Stimulation intensities and pulse characteristics were randomized across subjects to eliminate order effects.

Of note, some prior studies have used intracranial electrodes for mouse ECS (Rosenthal et al., 2025). This method relies on conclusions drawn from awake rats models and mismatched stimulation parameters, severely confounding its translational validity (Theilmann et al., 2014). In contrast, auricular ECS as used not only matches the non-invasiveness of patient ECT but also triggered patient-like phase III oscillations (Figure 1).

Electroconvulsive therapy (ECT) in humans

The retrospective treatment data of 28 random patients (10 male, 18 female, all > 18 years old), who underwent ECT at the Zentrum für Affektive, Stress-und Schlafstörungen (ZASS), University Psychiatric Clinics (UPK), Basel due to a psychiatric indication were included. Patients signed a general consent form (approved by the local ethics committee). All patient procedures were approved by and carried out in accordance with the Ethikkommission Nordwestund Zentralschweiz (EKNZ). One randomly chosen treatment within the first half of the ECT acute treatment series (i.e. < 7th treatment) per person was considered for analysis. Restimulation treatment data were excluded. Medical indication for ECT, patient information, treatment preparation and anesthesia followed standardized clinical protocols according to current standards of care. ECT was conducted with bifrontal stimulation using the Thymatron® System IV device (Somatics, LLC. Lake Bluff, IL, USA) and disposable adherent stimulus electrodes (EPAD, Thymapad®), as previously described (American Psychiatric Association, 2025), following Kellner’s ECT guidelines (Kellner, 2018). Briefly, patients were anesthetized with either propofol or etomidate (Braun Medical AG) and myorelaxed with succinylcholine chloride (Amino AG). Skin surface was cleansed using an alcohol-containing solvent and abrasive gel. A conductive and adhesive liquid (Signaspray®, Parker Laboratories) was then applied to the skin. EEG was recorded at 200 Hz using bilateral frontomastoid placement and filtered using a band-pass with 3dB points at 2 and 25 Hz. We excluded the first two seconds of recording following the stimulation to remove a capacitive coupling artifact. ECT was performed about 4 min after the injection of etomidate and 5 min after the injection of propofol, using the ‘double dose’ stimulation protocol. ECT sessions included in the analysis were performed by multiple clinicians with standardized preparation and stimulation procedures. Data were measured using the Thymatron during each session, including stimulation parameters and EEG derived seizure quality parameters. In addition, we had access to formalized clinical documentation listing the administered medication and manually assessed seizure duration. Digitalization of stored ECT data for the current analysis was performed after December 2020 using GENET-GPD (Freundlieb et al., 2023). The digitalized treatment data were derived from the GENET-GPD database. Patient data were stored in the UPK patient case files in compliance with accepted data protection and record-keeping regulations.

Electroencephalography (EEG) in mice

Mice were anesthetized as described above, and the glass or polymer window was removed. An EEG multielectrode array with 30 recording sites (EEG mouse-30-A-10-H32, Neuronexus) was used to record neural activity across the surface of the dorsal cortex. Signals were amplified using a 32 channel headstage (RHD 32ch, Intan Technologies) and digitized at 30 kHz using an Open Ephys acquisition board (Siegle et al., 2017). During ECS delivery, the EEG array was physically disconnected by an electronically triggered relay to prevent electrical arcing between electrodes. EEG data was downsampled to 1 kHz and band-pass filtered between 0.5 and 10 Hz. We excluded the first two seconds of recording following EEG reconnection to remove a capacitive coupling artifact (Figure S1). As the concurrent widefield calcium signal provided an unambiguous polarity reference, the polarity of the EEG signal is such that positive voltage deflections corresponded to increases in fluorescence.

Widefield imaging

Widefield imaging experiments were conducted on a custom-built macroscope with objectives mounted face-to-face (Nikon 85 mm/f1.8 sample side, Nikon 50 mm/f1.4 sensor side). A 470 nm LED (Thorlabs) powered by a custom-built LED driver was used to excite fluorescent indicators through an excitation filter (SP490, Thorlabs) reflected off of a dichroic mirror (LP490, Thorlabs) placed parfocal to the objectives. The fluorescence was collected through a 525/50 emission filter on a sCMOS camera (PCO edge 4.2). LED illumination was adjusted with a collimator (Thorlabs SM2F32-A) to homogenously illuminate cortical surface through the cranial window. Images were acquired at 100 Hz frame rate. Raw images were cropped on the sensor and data was stored to disk with custom-written LabVIEW (National Instruments) software, resulting in an effective pixel size of 60 μm2 at a resolution of 1108 by 1220 pixel (1.35 MP).

Two-photon imaging

Two-photon calcium imaging was performed using custom-built microscopes (Leinweber et al. 2014). The illumination source was a tunable femtosecond laser (Insight, Spectra Physics or Chameleon, Coherent) tuned to 930 nm. Emission light was band-pass filtered using a 525/50 filter and detected using a GaAsP photomultiplier (H7422, Hamamatsu). Photomultiplier signals were amplified (DHPCA-100, Femto), digitized (NI5772, National Instruments) at 800 MHz, and band-pass filtered at 80 MHz using a digital Fourier-transform filter implemented in custom-written software on an FPGA (NI5772, National Instruments). The scanning system of the microscopes was based on a 12 kHz resonant scanner (Cambridge Technology). Images were acquired at a resolution of 750 by 400 pixels (60 Hz frame rate) with a 16x, 0.8 NA water immersion objective (Nikon). We imaged layer 2/3 at a depth of 100-250 μm with a field of view of 375 by 300 µm. For prism imaging, to enable imaging of as much cortical depth as possible, the field of view was expanded to 270 by 960 µm, at a resolution of 750 by 800 pixels. This resulted in an effective frame rate of 30 Hz. Laminar boundaries were estimated from the density of GCaMP6-expressing cells, which typically peaks in layer 2/3 and upper layer 5 (Andermann et al., 2013).

Image processing

For analysis of chronic widefield macroscope imaging, raw movie data was manually registered across days by aligning subsequent mean projections of the data to the first recorded image sequence. In the raw 1108 by 1220 pixel images, we tiled the cortex with 818 ROIs each with a fixed location relative to bregma and lambda. This resulted in an average ROI size of 25 by 25 pixels (approximately 230 by 230 μm). Each ROI was also registered against the Allen Institute Mouse Brain Coordinate Framework (Wang et al., 2020). In concurrent EEG and widefield recordings, we defined annular ROIs centered on each of the 30 electrode sites. The inner boundary corresponded to the electrode footprint with a 50 μm buffer (approximately 60 pixels in diameter; approximately 550 μm) and was excluded from analysis, while the outer boundary defined a circular region with a diameter of approximately 80 pixels. Two-photon calcium imaging data were processed as previously described (Attinger et al., 2017; Keller et al., 2012). Briefly, raw images were full-frame registered to correct for lateral brain motion. Somata ROIs were manually selected based on mean and maximum fluorescence images. For each somata ROI, a corresponding neuropil ROI of equal pixel area was created as a surrounding annulus, maintaining a 3-pixel exclusion buffer and avoiding overlap with other defined ROIs. No neuropil subtraction to somata activity was done. Activity was calculated as the ΔF/F0, where F0 was the median fluorescence of the recording before ECS (typically 30 seconds). ECS recordings lasted for 5 min.

Histology

Mice underwent a single ECS or sham procedure while being monitored for calcium events through widefield imaging and were returned to their home cage for one hour. A subset (11/17) of mice was instead kept head-fixed and imaged for 1 hour to verify that no additional calcium events occurred. Mice were then anesthetized with a mixture of fentanyl (0.05 mg/kg; Actavis), midazolam (5.0 mg/kg; Dormicum, Roche), and medetomidine (0.5 mg/kg; Domitor, Orion), and transcardially perfused with phosphate-buffered saline (PBS) solution and subsequently with ice-cold 4% paraformaldehyde (PFA) in PBS. Brains were extracted and post-fixed overnight in 4% PFA at 4 °C. They were then embedded in 4% agarose and coronally sectioned through the entire anteroposterior extent of the cortex at (50 µm thickness) using a vibratome (Leica VT100S). We collected every fourth section for further processing, resulting in an effective sampling step of 200 µm. Free-floating sections were then processed for Fos immunohistochemistry. Briefly, sections were blocked for 2 h at room temperature in a solution containing 5% donkey serum (BSA, Sigma-Aldrich) and 0.5% Triton X-100 in PBS, then incubated overnight at 4 °C with a solution containing anti-Fos antibody (226008, Synaptic Systems, 1:1000), 1% Donkey Serum and 0.1% Triton X-100 in PBS. Wells were then rinsed 3 times in PBS (10 min each) and incubated with a solution containing donkey polyclonal anti-rabbit antibody (A31573, Invitrogen, Alexa Fluor™ 647 conjugate, 1:300), 1% Donkey Serum and 0.1% Triton X-100 in PBS for 2 h at room temperature. Following additional 10-minutes PBS washes, nuclei were counterstained with DAPI (Sigma-Aldrich, 1:1000) for 10 minutes. Finally, sections were mounted on microscopy slides with a mounting medium (VECTASHIELD, H-100, Vector Laboratories), sealed with nail polish, and stored at 4°C. Images were then acquired with an Axioscan Z1 (Zeiss) microscope using an apochromatic 10x/0.5 M27 objective (Zeiss). Fluorophores were excited with 422 nm and 668 nm X-Cite Xylis LED. The emission wavelength were filtered with a BP 445/50 filter for DAPI and a BP690/50 filter for Alexa 647. Images were collected with an Hamamatsu ORCA flash 4.0V camera at a pixel size of 0.650 µm.

Tissue registration and quantification

Images were aligned to the Allen Mouse Brain Common Coordinate Framework (CCFv3-2017) as described previously (Lupori et al., 2023). All preprocessing and registration steps, unless stated otherwise, were carried out using custom-written MATLAB (MathWorks) code. For each brain, images were ordered along the posterior-to-anterior direction. Based on a reference mark identifying the right side and/or tissue irregularities, images were mirrored vertically so that all matching hemispheres would be on the same side. Binary masks to restrict quantification to brain tissue were automatically generated from the DAPI images by training a pixel classifier in Ilastik (Berg et al. 2019). Each mask was then visually inspected and manually adjusted based on the Fos images, to remove regions containing tissue damage or staining artifacts. Left-right binary masks defining which hemisphere a pixel belongs to were then generated by manually drawing a segment at the midline of each slice. Using the DAPI images, each slice was matched to a plane in the CCFv3 though one first step of global alignment using QuickNII v2.2 (Puchades et al. 2019) and a second step of local non-rigid transformation with VisuAlign v0.9 (RRID:SCR_017978, VisuAlign). For cell counting, Fos channel images were first pre-processed with a rolling ball filter (50 pixels or 32.5 μm radius) for background removal. Cell detection and counting were then performed on the filtered images with custom scripting inside QuPath v.0.6.0 (Bankhead et al. 2017). Briefly, we used the inbuilt automatic cell detection tool to identify Fos-positive cells using the median fluorescence of the slide as a threshold. A random forest classifier was then trained to remove detection artifacts. For each cell, we calculated mean pixel intensity and x-y coordinates of the centroid. For all regions descending from isocortex, as per the hierarchy defined in the CCFv3, we computed cell density (number of cells/total area in mm2), mean cell intensity, and energy (cell density*mean cell intensity) (Lein et al., 2007; Lupori et al., 2023). For each metric, the value was then normalized by the average of that metric across all sham-treated mice.

Data analysis

All data analysis was performed with custom-written MATLAB (MathWorks) code.

When visualizing data with a box plot, the central box indicates the interquartile range (IQR), spanning from the first (Q1) to the third (Q3) quartiles. Whiskers extend to data points not considered outliers. Outliers were defined as data points lying outside of the Q1 – 1.5 IQR to Q3 + 1.5 IQR range. A central horizontal line indicates the median and an open circle the mean of the distribution. To represent the underlying probability density, a Kernel Density Estimate was computed using a Parzen-window method with a normally distributed kernel.

In Figure 1 and Figure 6, spectrograms were computed using a 2 s sliding window with 95 % overlap of the band-passed signal between 0.5 and 10 Hz. Power spectral density was estimated for each channel or ROI independently and converted to decibel scale. Oscillations were detected in the 1-5 Hz band by computing the Hilbert envelope of the EEG signal. Oscillation onsets were defined as the first crossing of the 90th percentile of the envelope after ECS. Oscillation terminations were defined as the first time the envelope dropped below a baseline threshold. Patient EEG is interrupted rapidly after the termination of oscillations. To account for the differences in recording length, the termination threshold was set as the 50th percentile Hilbert envelope for mice and as the 5th percentile for patients. For patient EEG, this detection method yielded oscillation durations not statistically different to those computed by the proprietary Thymatron algorithm (hierarchical bootstrap with resampling, p = 0.268). Power-weighted mean frequencies were computed as the sum of power values in a frequency, multiplied by that frequency. We reported the median values in a 2 s window centered on the global frequency peak and in the final 2 s of the detected oscillation. For mouse widefield data, the onsets and offsets time were computed from the concurrent EEG signal. We excluded frequencies between 7 and 9 Hz, if they exceeded 90 μV; these regions were dilated by 2 s and masked-out from further analysis. We completely excluded EEG electrodes and their associated widefield ROI if the onset/offset values were non-finite or if the total duration of oscillation was less than 1s, resulting in the exclusion of 78/480 (16%) of electrodes in our dataset.

In Figure 2 and Figure S3, calcium event speed was estimated through the local fluorescence gradient. Peak latencies were defined as the global maximum occurring 10 to 200 seconds after local wavefront onset. For each ROI, raw fluorescence values were low-passed filtered below 0.5 Hz and a plane (t = Sxx + Syy + t0) was fitted to the peak latencies of neighbors within a 0.8 mm radius using bisquare weighting. This determined a local slowness vector, the magnitude of which was inverted to determine speed. Speed values deviating more than 3.5 median absolute deviation from the session median were excluded. For two-photon imaging, propagation speed was calculated as the median velocity of calcium events across pairs of somata ROIs separated by at least 200 µm.

In Figure 2 and Figure S4, the period of direct ECS stimulation is defined as the first 5 s post ECS onset, based on the duration reported in Figure S1.

In Figure 2, 3 and Figure S3, the onset of a calcium event was defined as the first crossing of 150% the mean calcium activity before ECS. Only events occurring 10 s after ECS were considered as calcium events. For analysis of time to peak, we excluded ROIs for which there was no clear separation between direct ECS response and calcium event. To determine the presence of an oscillation, we averaged the calcium traces across hemispheres and used the same algorithm used in Figure 1 and Figure 6. To characterize the probability of calcium event or oscillation as a function of ECS parameters, we fitted a linear slope with ordinary least squares method. 95% confidence bands for the mean predicted response were computed across the range of predictor variables.

In Figure 2, to generate isochrones, we set a ROI-wise threshold to 150% of the maximum ECS baseline activity. The first time at which each ROI crossed this threshold was used to construct a time-of-activation map, which was then filtered with a normal distribution. Contour lines were extracted at 2 second intervals from the onset of calcium event until 50 s post-onset.

In Figure 4, cortical layers were separated in supragranular and infragranular, based on empirical distribution of cell densities, which shows a clear gap in cortical layer 4. To compare the activity of somata with their neuropil, we aligned each somata with the peak of its neuropil during calcium events. This alignment ensures that the propagation delay across the field of view is not taken into account. To compare supragranular and infragranular layers, somata were instead aligned to the max of a neuropil trace, averaged across all neuropil ROIs in the entire recording. To quantify the dynamics of GCaMP6f fluorescence following the calcium event, we computed a linear recovery slope with least-square ordinary fit in the 2 to 60s seconds following peak somata signal.

In Figure 6, the mid-ictal amplitude was computed as the mean of all oscillation peaks in a 4 s window centered at the duration of the oscillation (Krystal et al., 1995). The postictal suppression index was defined as the ratio of the median 1-30 Hz envelope amplitude post-oscillation over intra-oscillations. Welch’s periodogram method was used to compute the inter-hemispheric coherence between pairs of symmetric electrodes. We reported the values of coherence in the 1 to 5 Hz band.

In Figure S2, to illustrate the synchronicity of EEG and calcium signals, oscillation peaks in the EEG channel were detected using the Automatic Multiscale Peak Detection algorithm, which is suited to non-stationary signals, such as EEG, without requiring manual parameter tuning or prefiltering (Scholkmann et al., 2012).

Statistical analysis

All statistical information for the tests performed in the manuscript is provided in Table S1. Analyses accounted for the nested structure of the data consisting of recording units (neurons, EEG channels or ROIs) within subjects (mice or patients) using hierarchical bootstrap (Saravanan et al., 2020). Briefly, we first resampled with replacement data at the subject level and then resampled with replacement a second time at the recording unit level. We then computed the mean of this bootstrap sample and repeated this N times to generate a bootstrap distribution of the mean estimate. For all statistical testing the number of bootstrap samples (N) was 10 000, for plotting bootstrap mean and standard error response curves it was 1000. The bootstrap standard error is the 68% confidence interval (1 SD, 15.8th percentile to 84.2nd percentile) in the bootstrap distribution of the mean.

To account for the false discovery rate in multiple comparisons, we applied an adjustment to p-values (Benjamini and Hochberg, 1995). Briefly, p-values were sorted in ascending order and multiplied by the ratio of the total number of comparisons within a data group (for example, all bootstrap tests for a specific EEG metric) to their sorting rank. The final significance cutoff was determined by finding the largest p-value smaller than its rank-specific threshold, at which point all p-values with a higher rank (i.e., those with smaller p-values) were also rejected.

For linear regression analysis, we computed the F-statistic to test the significance of the model. This test evaluates whether the slope is significantly different from zero by comparing the model degrees of freedom (the number of predictors, here, one) against the degrees of freedom for error (the number of observations minus the parameters for the intercept and slope). The resulting F-statistic represents the ratio of explained variance to unexplained variance, quantifying the portion of the data’s variability accounted for by the linear regression. For multivariate linear regression analysis, we measured the probability of occurrence of a calcium event as a function of stimulation amplitude, frequency, and their two-way interaction. For each of those three parameters, we performed a two-sided t-test against the null hypothesis that this parameter is equal to zero.